---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.alt = "ggchangepoint plot of a time series with detected changepoints"
)
```

# ggchangepoint

<!-- badges: start -->
[![R-CMD-check](https://github.com/PursuitOfDataScience/ggchangepoint/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PursuitOfDataScience/ggchangepoint/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ggchangepoint)](https://CRAN.R-project.org/package=ggchangepoint)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

ggchangepoint provides a unified tidy interface to changepoint analysis in R. It wraps
multiple detection engines (changepoint, changepoint.np, ecp, wbs, breakfast, not, mosum,
fpop, IDetect) behind a consistent S3 result class (`ggcpt`) with `broom`-style
methods (`tidy()`, `glance()`, `augment()`), `ggplot2` integration via `autoplot()` and
custom geoms, and a full method-comparison, evaluation, simulation, and visualisation
toolkit.

The engines beyond `changepoint`, `changepoint.np`, and `ecp` are optional
(`Suggests`); install the ones you need. The original 0.1.0 functions
(`cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, `ggecpplot()`) continue to work
unchanged.

## Installation

Install the released version from CRAN:

``` r
install.packages("ggchangepoint")
```

Or the development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("PursuitOfDataScience/ggchangepoint")
```

## Quick start

```{r}
library(ggchangepoint)
library(ggplot2)
```

Generate a series with a mean shift:

```{r}
set.seed(2022)
x <- c(rnorm(100, 0, 1), rnorm(100, 10, 1))
```

Detect changepoints with the unified `cpt_detect()`:

```{r}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
res
```

The result is a `ggcpt` S3 object. Print it to see the changepoints, or use `tidy()`,
`glance()`, and `augment()`:

```{r}
tidy(res)
glance(res)
```

Visualise with `autoplot()`:

```{r}
autoplot(res)
```

## Unified detection across engines

`cpt_detect()` dispatches to any supported method by name:

```{r}
cpt_detect(x, method = "binseg", change_in = "mean")
cpt_detect(x, method = "wbs", change_in = "mean")
cpt_detect(x, method = "fpop", change_in = "mean")
```

Use `cpt_methods()` to see all available and planned methods with their engine
packages and installation status:

```{r}
cpt_methods()
```

## Compare methods

```{r}
ggcpt_compare(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

For a numeric summary, use `ggcpt_compare_table()`:

```{r}
ggcpt_compare_table(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

## Evaluation

When ground truth changepoints are known, compute accuracy metrics:

```{r}
cpt_metrics(pred = c(100), truth = c(100), n = 200)
```

## Data simulation

```{r}
dat <- cpt_simulate(200, changepoints = c(100), change_in = "mean",
                    params = c(0, 10), sd = 1)
attributes(dat)$true_changepoints
```

An alias `rcpt()` is provided for compatibility. Built-in test signals include
`signal_blocks()`, `signal_fms()`, `signal_mix()`, `signal_teeth()`,
and `signal_stairs()`.

## Penalty configuration

Use `cpt_penalty()` to construct penalty values for use with detection methods:

```{r}
cpt_penalty("BIC", n = 200)
cpt_penalty("AIC", n = 200)
cpt_penalty("Manual", value = 10)
```

## Direct engine wrappers

For fine-grained control, each detection engine has its own wrapper that returns
a `ggcpt` object directly:

```{r}
fpop_wrapper(x, penalty = 2 * log(200))
wbs_wrapper(x, n_intervals = 2000)
wbs2_wrapper(x)
not_wrapper(x, contrast = "pcwsConstMean")
mosum_wrapper(x)
idetect_wrapper(x)
tguh_wrapper(x)
```

## Custom geoms, stats, and theming

The package provides composable ggplot2 layers for changepoint visualisation:

```{r, eval = FALSE}
library(ggplot2)

# Use geom_changepoint as a standalone layer
cp_tbl <- tidy(cpt_detect(x, method = "pelt", change_in = "mean"))
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_changepoint(data = cp_tbl, aes(xintercept = cp), color = "red") +
  theme_ggcpt()

# Use stat_changepoint to compute and draw changepoints in one step
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  stat_changepoint(method = "pelt", color = "red")

# Shade alternating segments between changepoints
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  annotate_segments(cp = cp_tbl$cp, n = length(x))

# Highlight segments with geom_cpt_segment
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_cpt_segment(data = cp_tbl, aes(xintercept = cp), color = "blue")

# Draw confidence intervals with geom_cpt_ci (when the engine provides them)
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_cpt_ci(data = cp_tbl, aes(xintercept = cp, ymin = lower, ymax = upper))
```

## Evaluation metrics with annotation

When multiple annotation sets are available, use `cpt_metrics_annotated()` and
visualise with `ggcpt_eval()`:

```{r}
cpt_metrics_annotated(c(100), list(c(100), c(101), c(99)), n = 200, margin = 5)
```

## Class constructors

Advanced users can construct `ggcpt` objects directly or test for the class:

```{r, eval = FALSE}
new_ggcpt(
  changepoints = tibble::tibble(cp = 100L, cp_value = 5.0),
  data = tibble::tibble(index = 1:200, value = rnorm(200)),
  method = "manual"
)
is_ggcpt(x)
```

## Original ecp wrapper

The `ecp_wrapper()` and its plotting function `ggecpplot()` provide direct
access to the ecp engine:

```{r, eval = FALSE}
ecp_wrapper(x, algorithm = "divisive")
ggecpplot(x, algorithm = "divisive")
```

## Original wrappers (0.1.0 API)

The original `cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, and `ggecpplot()`
continue to work unchanged for backward compatibility.

```{r}
cpt_wrapper(x)
ggcptplot(x)
```

## Additional S3 methods

The `ggcpt` class also provides:

```{r}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
summary(res)          # human-readable digest
as_tibble(res)        # tibble of changepoints
as.data.frame(res)    # data frame of changepoints
format(res)           # one-line summary string
plot(res)             # base-graphics fallback (delegates to autoplot)
```

## Learn more

See the vignettes for a comprehensive walkthrough:

- `vignette("ggchangepoint", package = "ggchangepoint")` — feature tour
- `vignette("introduction", package = "ggchangepoint")`
- `vignette("comparison", package = "ggchangepoint")`
