---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.alt = "ggchangepoint plot of a time series with detected changepoints"
)
```

# ggchangepoint

<!-- badges: start -->
[![R-CMD-check](https://github.com/PursuitOfDataScience/ggchangepoint/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PursuitOfDataScience/ggchangepoint/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ggchangepoint)](https://CRAN.R-project.org/package=ggchangepoint)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

ggchangepoint provides a unified tidy interface to changepoint analysis in R.
One dispatcher — `cpt_detect()` — covers **31 detection methods across five
methodological families**, every result comes back as the same tidy `ggcpt`
S3 object with `broom`-style methods (`tidy()`, `glance()`, `augment()`), and
every result renders directly with `ggplot2` via `autoplot()` and a family of
composable geoms.

The five families (run `cpt_methods()` for the live table):

- **Penalised / optimal partitioning** — PELT, BinSeg, SegNeigh, AMOC
  (`changepoint`), FPOP (`fpop`), the CROPS penalty path (`cpt_crops()`),
  fastcpd incl. AR/ARMA/GARCH model changes (`fastcpd`), and exact
  change-in-slope via CPOP (`cpop`).
- **Multiscale / search** — WBS (`wbs`), WBS2 and TGUH (`breakfast`), NOT
  (`not`), MOSUM incl. multiscale (`mosum`), Isolate-Detect (`IDetect`), and
  SMUCE/HSMUCE with changepoint-location **confidence intervals** (`stepR`).
- **Nonparametric / kernel** — ED-PELT (`changepoint.np`),
  E-Divisive/E-Agglo (`ecp`), kernel running statistics (`kcpRS`), NP-MOJO
  (`CptNonPar`), sequential CPM (`cpm`), self-normalisation (`SNSeg`).
- **Bayesian** — Barry–Hartigan posterior probabilities (`bcp`), online
  BOCPD with run-length posteriors (`ocp`), BEAST Bayesian model averaging
  (`Rbeast`).
- **Multivariate / high-dimensional / regression** — sparse-projection
  inspect (`InspectChangepoint`), online high-dimensional ocd (`ocd`),
  geometric mapping (`changepoint.geo`), Bai–Perron structural breaks with
  CIs (`strucchange`), broken-line regression (`segmented`),
  changepoint-vs-trend-vs-autocorrelation model selection (`EnvCpt`), and
  robust detection under drift and AR noise (`DeCAFS`).

Only `changepoint`, `changepoint.np`, and `ecp` are hard dependencies; every
other engine is optional (`Suggests`) — install the ones you need and
`cpt_methods()` will tell you what is missing. The original 0.1.0 functions
(`cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, `ggecpplot()`) continue to
work unchanged.

## Installation

Install the released version from CRAN:

``` r
install.packages("ggchangepoint")
```

Or the development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("PursuitOfDataScience/ggchangepoint")
```

## Quick start

```{r}
library(ggchangepoint)
library(ggplot2)
```

Generate a series with a mean shift:

```{r}
set.seed(2022)
x <- c(rnorm(100, 0, 1), rnorm(100, 10, 1))
```

Detect changepoints with the unified `cpt_detect()`:

```{r}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
res
```

The result is a `ggcpt` S3 object. Print it to see the changepoints, or use `tidy()`,
`glance()`, and `augment()`:

```{r}
tidy(res)
glance(res)
```

Visualise with `autoplot()`:

```{r}
autoplot(res)
```

## Unified detection across engines

`cpt_detect()` dispatches to any supported method by name:

```{r}
cpt_detect(x, method = "binseg", change_in = "mean")
cpt_detect(x, method = "wbs", change_in = "mean")
cpt_detect(x, method = "fpop", change_in = "mean")
```

Use `cpt_methods()` to see all available and planned methods with their engine
packages and installation status:

```{r}
cpt_methods()
```

## Detection with uncertainty: confidence intervals and posteriors

SMUCE (`smuce_wrapper()`, via `stepR`) delivers a confidence interval for
every changepoint location; Bai–Perron (`strucchange_wrapper()`) and
broken-line regression (`segmented_wrapper()`) do the same for regression
breaks. The intervals live in `ci_lower`/`ci_upper` columns and render with
`autoplot(show_ci = TRUE)` or the `geom_cpt_ci()` layer:

```{r, eval = requireNamespace("stepR", quietly = TRUE)}
res_smuce <- smuce_wrapper(x)
tidy(res_smuce)
autoplot(res_smuce, show_ci = TRUE, show_fit = TRUE)
```

Bayesian engines return posterior probabilities instead: `bcp_wrapper()`
(posterior probability of a change at every location), `beast_wrapper()`
(Bayesian model averaging), and `bocpd_wrapper()` (online run-length
posterior). Two dedicated displays accompany them — `ggcpt_posterior()` and
`ggcpt_runlength()`:

```{r, eval = requireNamespace("bcp", quietly = TRUE)}
res_bcp <- bcp_wrapper(x, seed = 1)
tidy(res_bcp)
ggcpt_posterior(res_bcp)
```

## The penalty path: CROPS

Instead of guessing one penalty, `cpt_crops()` computes **every** optimal
segmentation over a penalty range (Haynes, Eckley and Fearnhead, 2017) and
plots the elbow diagnostic, the penalty path, or the candidate segmentations
themselves:

```{r}
path <- cpt_crops(c(rnorm(100), rnorm(100, 3), rnorm(100, -1)))
path
autoplot(path)                          # cost elbow
```

```{r}
autoplot(path, type = "segmentations")  # see the actual candidate models
```

## Compare methods

```{r}
ggcpt_compare(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

For a numeric summary, use `ggcpt_compare_table()`:

```{r}
ggcpt_compare_table(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

## Batch detection and stability diagnostics

`cpt_batch()` runs one detector over many series (a matrix, data frame, or
list) and returns a tidy tibble of results — honouring `future::plan()` for
parallel execution. `cpt_stability()` bootstrap-resamples within fitted
segments and reports how often each location is re-detected, a cheap
confidence signal for engines with no native intervals:

```{r}
X <- cbind(shifted = x, noise = rnorm(200))
batch <- cpt_batch(X, method = "pelt")
batch
autoplot(batch)
```

```{r}
st <- cpt_stability(x, method = "pelt", B = 50, seed = 1)
st
autoplot(st)
```

## Multivariate and high-dimensional detection

The multivariate engines accept a matrix (rows are time points) directly
through `cpt_detect()` and render as faceted small-multiples:

```{r, eval = requireNamespace("InspectChangepoint", quietly = TRUE)}
set.seed(1)
Xhd <- cbind(a = c(rnorm(80), rnorm(80, 3)),
             b = c(rnorm(80), rnorm(80, -2)),
             c = rnorm(160))
res_hd <- inspect_wrapper(Xhd)
tidy(res_hd)
autoplot(res_hd)
```

## Evaluation

When ground truth changepoints are known, compute accuracy metrics
(precision/recall/F1 under one-to-one matching, the covering metric,
Hausdorff distance, adjusted Rand index):

```{r}
cpt_metrics(pred = c(100), truth = c(100), n = 200)
```

When multiple annotation sets are available, use `cpt_metrics_annotated()`,
and visualise agreement with `ggcpt_eval()`:

```{r}
cpt_metrics_annotated(c(100), list(c(100), c(101), c(99)), n = 200, margin = 5)
```

## Data simulation

```{r}
dat <- cpt_simulate(200, changepoints = c(100), change_in = "mean",
                    params = c(0, 10), sd = 1)
attributes(dat)$true_changepoints
```

An alias `rcpt()` is provided for compatibility. Built-in test signals include
`signal_blocks()` (the Donoho–Johnstone blocks signal), `signal_fms()`,
`signal_mix()`, `signal_teeth()`, and `signal_stairs()`.

## Penalty configuration

Use `cpt_penalty()` to construct penalty values for use with detection methods:

```{r}
cpt_penalty("BIC", n = 200)
cpt_penalty("AIC", n = 200)
cpt_penalty("Manual", value = 10)
```

## Direct engine wrappers

For fine-grained control, every engine has its own wrapper returning a
`ggcpt` object directly. The classic search/pruning engines:

```{r}
fpop_wrapper(x, penalty = 2 * log(200))
wbs_wrapper(x, n_intervals = 2000)
wbs2_wrapper(x)
not_wrapper(x, contrast = "pcwsConstMean")
mosum_wrapper(x)
idetect_wrapper(x)
tguh_wrapper(x)
```

And the 0.4.0 wave (each behind its `Suggests` engine): `smuce_wrapper()`,
`cpop_wrapper()`, `bcp_wrapper()`, `bocpd_wrapper()`, `beast_wrapper()`,
`cpm_wrapper()`, `kcp_wrapper()`, `npmojo_wrapper()`, `decafs_wrapper()`,
`sn_wrapper()`, `inspect_wrapper()`, `ocd_wrapper()`, `geomcp_wrapper()`,
`strucchange_wrapper()`, `segmented_wrapper()`, `envcpt_wrapper()`, and
`fastcpd_wrapper()`.

```{r, eval = requireNamespace("cpop", quietly = TRUE)}
# change-in-slope: exact penalised broken-line estimation
y_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200)
res_slope <- cpop_wrapper(y_slope)
autoplot(res_slope, show_fit = TRUE)
```

## Custom geoms, stats, and theming

The package provides composable ggplot2 layers for changepoint visualisation:

```{r, eval = FALSE}
library(ggplot2)

# Use geom_changepoint as a standalone layer
cp_tbl <- tidy(cpt_detect(x, method = "pelt", change_in = "mean"))
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_changepoint(data = cp_tbl, aes(xintercept = cp), color = "red") +
  theme_ggcpt()

# Use stat_changepoint to compute and draw changepoints in one step
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  stat_changepoint(method = "pelt", color = "red")

# Shade alternating segments between changepoints
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  annotate_segments(cp = cp_tbl$cp, n = length(x))

# Highlight segments with geom_cpt_segment
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_cpt_segment(data = cp_tbl, aes(xintercept = cp), color = "blue")

# Draw confidence intervals with geom_cpt_ci (when the engine provides them)
cp_ci <- tidy(smuce_wrapper(x))
ggplot(data.frame(index = seq_along(x), value = x), aes(index, value)) +
  geom_line() +
  geom_cpt_ci(data = cp_ci,
              aes(x = cp, xmin = ci_lower, xmax = ci_upper, y = min(x) - 1))
```

## Interactive exploration and citations

Any result (or ggplot built from one) renders as an interactive widget with
`ggcpt_interactive()` (requires `plotly`). And `cpt_cite()` returns the
methodological reference behind a result, so analyses can cite the right
paper without leaving R:

```{r}
cpt_cite("pelt")
```

```{r, eval = FALSE}
ggcpt_interactive(res)   # hover for values; requires plotly
```

## Class constructors

Advanced users can construct `ggcpt` objects directly or test for the class:

```{r, eval = FALSE}
new_ggcpt(
  changepoints = tibble::tibble(cp = 100L, cp_value = 5.0),
  data = tibble::tibble(index = 1:200, value = rnorm(200)),
  method = "manual"
)
is_ggcpt(x)
```

## Original ecp wrapper

The `ecp_wrapper()` and its plotting function `ggecpplot()` provide direct
access to the ecp engine (including genuine multivariate input):

```{r, eval = FALSE}
ecp_wrapper(x, algorithm = "divisive")
ggecpplot(x, algorithm = "divisive")
```

## Original wrappers (0.1.0 API)

The original `cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, and `ggecpplot()`
continue to work unchanged for backward compatibility.

```{r}
cpt_wrapper(x)
ggcptplot(x)
```

## Additional S3 methods

The `ggcpt` class also provides:

```{r}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
summary(res)          # human-readable digest
as_tibble(res)        # tibble of changepoints
as.data.frame(res)    # data frame of changepoints
format(res)           # one-line summary string
plot(res)             # base-graphics fallback (delegates to autoplot)
```

## Learn more

See the vignettes for a comprehensive walkthrough:

- `vignette("ggchangepoint", package = "ggchangepoint")` — feature tour
- `vignette("introduction", package = "ggchangepoint")` — the framework, in
  research-paper form
- `vignette("comparison", package = "ggchangepoint")` — method comparison
  and evaluation
