---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggchangepoint

<!-- badges: start -->
[![R-CMD-check](https://github.com/PursuitOfDataScience/ggchangepoint/workflows/R-CMD-check/badge.svg)](https://github.com/PursuitOfDataScience/ggchangepoint/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/ggchangepoint)](https://CRAN.R-project.org/package=ggchangepoint)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

ggchangepoint provides a unified tidy interface to changepoint analysis in R. It wraps
multiple detection engines (changepoint, changepoint.np, ecp, wbs, breakfast, not, mosum,
fpop, IDetect, and gfpop) behind a consistent S3 result class (`ggcpt`) with `broom`-style
methods (`tidy()`, `glance()`, `augment()`), `ggplot2` integration via `autoplot()` and
custom geoms, and a full method-comparison, evaluation, and simulation toolkit.

The engines beyond `changepoint`, `changepoint.np`, and `ecp` are optional
(`Suggests`); install the ones you need. The original 0.1.0 functions
(`cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, `ggecpplot()`) continue to work
unchanged.

## Installation

Install the released version from CRAN:

``` r
install.packages("ggchangepoint")
```

Or the development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("PursuitOfDataScience/ggchangepoint")
```

## Quick start

```{r}
library(ggchangepoint)
library(ggplot2)
```

Generate a series with a mean shift:

```{r}
set.seed(2022)
x <- c(rnorm(100, 0, 1), rnorm(100, 10, 1))
```

Detect changepoints with the unified `cpt_detect()`:

```{r}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
res
```

The result is a `ggcpt` S3 object. Print it to see the changepoints, or use `tidy()`,
`glance()`, and `augment()`:

```{r}
tidy(res)
glance(res)
```

Visualise with `autoplot()`:

```{r}
autoplot(res)
```

## Unified detection across engines

`cpt_detect()` dispatches to any supported method by name:

```{r}
cpt_detect(x, method = "binseg", change_in = "mean")
cpt_detect(x, method = "wbs", change_in = "mean")
cpt_detect(x, method = "fpop", change_in = "mean")
```

## Compare methods

```{r}
ggcpt_compare(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

For a numeric summary, use `ggcpt_compare_table()`:

```{r}
ggcpt_compare_table(x, methods = c("pelt", "binseg", "fpop", "wbs"))
```

## Evaluation

When ground truth changepoints are known, compute accuracy metrics:

```{r}
cpt_metrics(pred = c(100), truth = c(100), n = 200)
```

## Data simulation

```{r}
dat <- cpt_simulate(200, changepoints = c(100), change_in = "mean",
                    params = c(0, 10), sd = 1)
attributes(dat)$true_changepoints
```

Built-in test signals include `signal_blocks()`, `signal_fms()`, `signal_mix()`,
`signal_teeth()`, and `signal_stairs()`.

## Original wrappers (0.1.0 API)

The original `cpt_wrapper()`, `ecp_wrapper()`, `ggcptplot()`, and `ggecpplot()`
continue to work unchanged for backward compatibility.

```{r}
cpt_wrapper(x)
ggcptplot(x)
```

## Learn more

See the vignettes for a comprehensive walkthrough:

- `vignette("introduction", package = "ggchangepoint")`
- `vignette("comparison", package = "ggchangepoint")`
