---
title: "Comparing and Evaluating Changepoint Methods with ggchangepoint"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing and Evaluating Changepoint Methods with ggchangepoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint plot comparing changepoint detection methods on a time series"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_light())

# Optional (Suggests) engines: gate the chunks that need them so the
# vignette builds on any installation.
has_fpop        <- requireNamespace("fpop", quietly = TRUE)
has_wbs         <- requireNamespace("wbs", quietly = TRUE)
has_stepR       <- requireNamespace("stepR", quietly = TRUE)
has_strucchange <- requireNamespace("strucchange", quietly = TRUE)

cmp_methods <- if (has_fpop && has_wbs) {
  c("pelt", "binseg", "fpop", "wbs")
} else {
  c("pelt", "binseg", "amoc")
}
```

# Abstract

No single changepoint detection method dominates across signal shapes, noise
regimes, and computational budgets: large-scale evaluations find that the
ranking of algorithms is highly dataset-dependent and that default
configurations are often far from optimal [@van2020evaluation;
@truong2020selective]. Sound practice therefore requires running several
detectors, comparing their outputs, and — when ground truth is available —
scoring them with well-defined accuracy metrics. This article presents the
comparison and evaluation toolkit of **ggchangepoint**: visual comparison
across methods (`ggcpt_compare()`), a common tidy representation of competing
segmentations (`ggcpt_compare_table()`), a metrics module implementing
precision/recall/F1 under one-to-one matching, the covering metric, Hausdorff
distance, and the adjusted Rand index (`cpt_metrics()`), multi-annotator
scoring in the style of the Turing Change Point Dataset benchmark
(`cpt_metrics_annotated()`), and visual evaluation against ground truth
(`ggcpt_eval()`). We further discuss three complementary tools for
quantifying the *uncertainty* of a segmentation — engine-native confidence
intervals, bootstrap stability profiles (`cpt_stability()`), and the CROPS
penalty path (`cpt_crops()`) — and close with a small simulation-based
benchmarking workflow built from the package's ground-truth generators.

# Introduction

The changepoint literature offers a wide menu of detectors — penalised
optimal partitioning, binary segmentation and its wild and
narrowest-over-threshold refinements, moving-sum statistics, nonparametric
divergence measures, Bayesian posteriors — each with its own inductive bias
[@aminikhanghahi2017survey; @truong2020selective]. Benchmarks that score many
algorithms on many series find no uniform winner: performance depends on the
kind of change (mean, variance, distribution), the noise (Gaussian, heavy
tailed, autocorrelated), the number and spacing of changes, and the tuning of
penalties and thresholds [@van2020evaluation]. Two practical consequences
follow. First, an analyst should *compare* several methods on the data at
hand rather than trust one default. Second, when ground truth (or expert
annotation) exists, comparison should be quantitative, using metrics with
agreed conventions.

**ggchangepoint** supports both activities behind one interface. Every
detector returns the same tidy `ggcpt` object, so competing methods are
directly comparable; the comparison module renders them side by side, and the
evaluation module scores them. This article is the package's methods-paper
treatment of that workflow. The companion vignette
`vignette("introduction", package = "ggchangepoint")` documents the full
detection surface; here we take detectors as given and focus on comparison,
evaluation, and uncertainty.

# Problem setup

Let $y_{1:n} = (y_1, \dots, y_n)$ be an ordered sequence. A segmentation with
$m$ changepoints is an ordered set $\tau_{0:m+1}$ with
$0 = \tau_0 < \tau_1 < \cdots < \tau_m < \tau_{m+1} = n$, partitioning the
index set into segments $A_j = \{\tau_{j-1}+1, \dots, \tau_j\}$,
$j = 1, \dots, m+1$. Most offline methods minimise a penalised cost
$$
\sum_{j=1}^{m+1} \mathcal{C}\bigl(y_{(\tau_{j-1}+1):\tau_j}\bigr) + \beta f(m),
$$
where $\mathcal{C}(\cdot)$ is a segment cost and $\beta f(m)$ guards against
over-segmentation [@yao1988estimating; @killick2012pelt].

Throughout the package a changepoint $\tau$ is reported in the **"left"
convention**: $\tau$ is the *last index of the left segment*, the convention
of the **changepoint** package [@killick2012pelt]. Engines that natively
report the first index of the right segment (e.g. the **ecp** family,
@matteson2014nonparametric) are normalised on the way in, so comparisons
across methods are always like for like.

Comparing segmentations is harder than comparing point estimates for three
reasons. First, the *number* of detected changes varies across methods, so a
metric must handle unequal-length sets. Second, a detection a few indices
away from a true change is usually acceptable, so metrics need a tolerance
margin — and a matching rule that prevents one true change from being
"claimed" by several detections. Third, changepoint sets induce
*partitions*, and two very different-looking point sets can induce similar
partitions; good practice therefore reports both point-based metrics
(precision/recall, Hausdorff) and partition-based metrics (covering, Rand)
[@van2020evaluation].

# Visual comparison

We simulate a series with three mean levels (two changepoints) and run
several detectors through the unified dispatcher. `ggcpt_compare()` accepts
the raw series and a vector of method names, runs `cpt_detect()` for each,
and renders the results. The default `layout = "facet"` draws one panel per
method:

```{r compare-facet}
set.seed(2024)
x <- c(rnorm(150, 0), rnorm(150, 3), rnorm(200, 1))
ggcpt_compare(x, methods = cmp_methods)
```

A method that finds *no* changepoints keeps its (empty-ruled) panel rather
than silently disappearing — a method that ran and found nothing is a result,
not a missing value:

```{r compare-nochange}
set.seed(7)
x_null <- rnorm(300)
ggcpt_compare(x_null, methods = c("pelt", "binseg"))
```

The `layout = "overlay"` variant superimposes all methods in a single panel
with colour-coded rules, which makes small disagreements in estimated
locations easier to see:

```{r compare-overlay}
ggcpt_compare(x, methods = cmp_methods, layout = "overlay")
```

For a numeric rather than visual comparison, `ggcpt_compare_table()` returns
one tidy tibble with a row per (method, changepoint) pair:

```{r compare-table}
ggcpt_compare_table(x, methods = cmp_methods)
```

Because every row carries the same columns, this table pipes directly into
**dplyr**/**ggplot2** summaries — counting detections per method, plotting
location agreement, and so on [@wickham2016ggplot2].

# Accuracy metrics

When the true changepoints $\mathcal{T} = \{t_1, \dots, t_K\}$ are known,
`cpt_metrics(pred, truth, n, margin)` scores a predicted set
$\mathcal{P} = \{p_1, \dots, p_M\}$ with the following quantities.

**Precision, recall, and F1 under one-to-one matching.** A prediction $p$
*matches* a truth $t$ if $|p - t| \le \texttt{margin}$. Matching is
one-to-one: predictions are scanned in increasing order and each takes the
earliest unmatched truth within the margin, so each truth is claimed by at
most one prediction (for points on a line this greedy rule attains a maximum
matching). With $\mathrm{TP}$ matched pairs,
$$
\mathrm{precision} = \frac{\mathrm{TP}}{|\mathcal{P}|}, \qquad
\mathrm{recall} = \frac{\mathrm{TP}}{|\mathcal{T}|}, \qquad
F_1 = \frac{2\,\mathrm{precision}\cdot\mathrm{recall}}
           {\mathrm{precision} + \mathrm{recall}},
$$
which guarantees all three lie in $[0, 1]$ — three predictions crowded around
one true change score one true positive, not three.

**The covering metric.** Following @van2020evaluation, let $\mathcal{S}$ and
$\mathcal{S}'$ be the partitions induced by the truth and the prediction, and
$J(A, A') = |A \cap A'| / |A \cup A'|$ the Jaccard index of two segments. The
covering of $\mathcal{S}$ by $\mathcal{S}'$ is
$$
\mathrm{cov}(\mathcal{S}, \mathcal{S}') =
  \frac{1}{n} \sum_{A \in \mathcal{S}} |A| \,
  \max_{A' \in \mathcal{S}'} J(A, A'),
$$
a weighted average of the best overlap achieved for each true segment.

**Hausdorff distance.** The worst-case location error
$\max\{\max_p \min_t |p - t|,\; \max_t \min_p |p - t|\}$; it is `NA` when
either set is empty (there is no distance to a nonexistent point).

**Adjusted Rand index.** The chance-corrected agreement of the two induced
segment labellings; 1 for identical partitions, 0 for chance-level
agreement.

**Annotation error and matched location errors.** The absolute difference in
counts $\bigl||\mathcal{P}| - |\mathcal{T}|\bigr|$, and the MAE/RMSE of the
matched location pairs.

```{r metrics-basic}
# perfect detection
cpt_metrics(pred = c(150, 300), truth = c(150, 300), n = 500)

# near misses within the margin still match one-to-one
cpt_metrics(pred = c(148, 305), truth = c(150, 300), n = 500, margin = 5)

# three predictions around one truth: one TP, precision 1/3
cpt_metrics(pred = c(148, 150, 152), truth = c(150), n = 500, margin = 5)
```

Three edge-case conventions deserve emphasis, because getting them wrong
silently corrupts benchmark averages:

- **Both sets empty.** Predicting "no changepoints" for a series with no
  changepoints is *exactly right*, so precision, recall, F1, covering, and
  the Rand index all equal 1.
- **Empty prediction, non-empty truth.** An empty changepoint set induces the
  trivial one-segment partition, which is a perfectly well-defined partition:
  covering scores it by segment overlap (not an automatic 0), while the
  adjusted Rand index is 0 (chance-level agreement) and precision/recall are
  0.
- **Out-of-range indices.** Changepoint locations must lie in
  $\{1, \dots, n-1\}$ under the left convention; anything outside is dropped
  with a warning rather than crashing the partition construction.

```{r metrics-edges}
# a correct "no change" answer is rewarded
cpt_metrics(pred = integer(0), truth = integer(0), n = 300)

# empty prediction against one true change: trivial-partition covering
cpt_metrics(pred = integer(0), truth = c(150), n = 300)
```

# Multi-annotator evaluation

For real data, "ground truth" is often a set of human annotations that
disagree with one another. The Turing Change Point Dataset benchmark
[@van2020evaluation] therefore scores a prediction against *each* annotator
and averages. `cpt_metrics_annotated()` implements this convention: it takes
a list of annotation vectors and returns the averaged precision, recall, F1,
and covering.

```{r annotated}
annotations <- list(
  ann1 = c(150, 300),
  ann2 = c(152, 301),
  ann3 = c(149)          # a third annotator missed the second change
)
cpt_metrics_annotated(pred = c(150, 300), annotations, n = 500, margin = 5)
```

# Visual evaluation

`ggcpt_eval()` overlays predictions and ground truth on the series, shades
the tolerance window around each true change, and colours each prediction as
a true positive or false positive, with missed truths drawn as dashed "Miss"
rules. It uses the *same one-to-one matching* as `cpt_metrics()`, so the
picture and the numbers always agree:

```{r eval-plot}
truth <- c(150, 300)
pred <- c(151, 240)      # one hit, one false alarm, one miss
ggcpt_eval(pred, truth, data_vec = x, margin = 5)
cpt_metrics(pred, truth, n = length(x), margin = 5)
```

# Uncertainty beyond point sets

A segmentation is a point estimate. Three complementary tools quantify how
much confidence it deserves.

## Engine-native confidence intervals

Some engines deliver genuine confidence statements for changepoint
*locations*. SMUCE [@frick2014smuce] controls, at level $\alpha$, the
probability of overestimating the number of changepoints, and returns a
confidence interval for every location; the Bai–Perron dynamic program
[@bai2003computation; @zeileis2002strucchange] returns break-date intervals
for regression breaks. Both populate `ci_lower`/`ci_upper` columns on the
`ggcpt` changepoints tibble, and `autoplot(show_ci = TRUE)` draws them as
whiskers:

```{r smuce-ci, eval = has_stepR}
res_smuce <- smuce_wrapper(x)
tidy(res_smuce)
autoplot(res_smuce, show_ci = TRUE, show_fit = TRUE)
```

```{r strucchange-ci, eval = has_strucchange}
res_bp <- strucchange_wrapper(x)
tidy(res_bp)
```

## Bootstrap stability

Most engines ship no intervals at all. `cpt_stability()` provides a cheap,
model-agnostic substitute: it fits the detector once, resamples residuals
*within* the fitted segments (preserving the estimated regime structure),
re-runs the detector on each replicate, and reports the proportion of
replicates that re-detect a change within `margin` of each index. Locations
that survive resampling are trustworthy; locations that appear in only a
fraction of replicates are fragile.

```{r stability}
st <- cpt_stability(x, method = "pelt", B = 30, seed = 1)
st
autoplot(st)
```

## Penalty-path sensitivity: CROPS

Penalised methods commit to one penalty $\beta$, and the segmentation can
change qualitatively as $\beta$ moves. CROPS [@haynes2017computationally]
computes *every* optimal segmentation as the penalty ranges over an interval,
at roughly one PELT run per distinct solution. `cpt_crops()` returns the full
path; the elbow plot shows where adding changepoints stops paying for itself,
and the segmentation facets show the actual candidate models being chosen
among — penalty selection as a diagnostic rather than a guess:

```{r crops}
path <- cpt_crops(x)
path
autoplot(path)                          # cost elbow
autoplot(path, type = "segmentations")  # the candidate segmentations
```

# A benchmarking workflow

The package's simulation module closes the loop: generators with *known*
truth feed the metrics module, so methods can be scored over replications.
`cpt_simulate()` draws series with specified changepoints, and the canonical
test signals ship as ready-made generators — including `signal_blocks()`, the
Donoho–Johnstone blocks signal [@donoho1994ideal], plus `signal_teeth()` and
`signal_stairs()`.

```{r signals, fig.height = 4}
blocks <- signal_blocks(n = 500, seed = 3)
ggplot(blocks, aes(index, value)) +
  geom_line(colour = "grey40") +
  geom_vline(xintercept = attr(blocks, "true_changepoints"),
             colour = "blue", linewidth = 0.3) +
  labs(title = "signal_blocks(): Donoho-Johnstone blocks with true changepoints")
```

A minimal benchmark: three methods, ten replications of a two-change series,
scored by precision, recall, F1, and covering. (For clarity we keep
everything sequential and small; `cpt_batch()` runs one method over many
series and honours `future::plan()` for parallel execution in larger
studies.)

```{r benchmark}
methods <- cmp_methods[1:3]
n_rep <- 10

results <- do.call(rbind, lapply(seq_len(n_rep), function(r) {
  dat <- cpt_simulate(400, changepoints = c(130, 260), change_in = "mean",
                      params = c(0, 2.5, 0.5), sd = 1, seed = 100 + r)
  truth <- attr(dat, "true_changepoints")
  do.call(rbind, lapply(methods, function(m) {
    fit <- cpt_detect(dat$value, method = m)
    score <- cpt_metrics(fit$changepoints$cp, truth, n = 400, margin = 5)
    cbind(tibble::tibble(rep = r, method = m),
          score[, c("precision", "recall", "f1", "covering")])
  }))
}))

# average over replications
res_summary <- aggregate(cbind(precision, recall, f1, covering) ~ method,
                         data = results, FUN = mean)
knitr::kable(res_summary, digits = 3,
             caption = "Mean accuracy over 10 replications (margin = 5).")
```

The same skeleton scales to the studies of @van2020evaluation: more
generators (heavy-tailed and autocorrelated noise via
`cpt_simulate(noise = "t")` and `noise = "ar1"`), more methods (everything in
`cpt_methods()`), more replications, and parallel execution via
`future::plan(multisession)`.

# Discussion

Comparison and evaluation are not afterthoughts in changepoint analysis; they
are how an analyst earns confidence in a segmentation. The design of
**ggchangepoint**'s toolkit follows three principles. First, *a common
representation makes comparison trivial*: because every engine returns the
same `ggcpt` contract in the same location convention, `ggcpt_compare()` and
the metrics module work for all thirty-plus methods without special cases.
Second, *metrics must agree with their pictures*: `ggcpt_eval()` and
`cpt_metrics()` share one matching routine, so a plotted true positive is a
counted true positive. Third, *uncertainty deserves first-class treatment*:
engine-native intervals, bootstrap stability, and penalty paths give three
independent views of how much a reported changepoint should be trusted, and
all three render directly with **ggplot2** [@wickham2016ggplot2].

For the detection surface itself — the dispatcher, the engine wave, the
Bayesian displays, and the multivariate tools — see
`vignette("introduction", package = "ggchangepoint")` and the feature tour in
`vignette("ggchangepoint", package = "ggchangepoint")`.

# References
