---
title: "Extending ggchangepoint"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending ggchangepoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 72,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint plot of a result from an externally supplied detector"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_ggcpt())
```

This package wraps fifty detectors. It will never wrap all of them,
and some of the most interesting ones it *cannot* wrap: `changeforest`
[@londschien2023changeforest] is on conda-forge only,
`ChangepointInference` [@jewell2022testing] is on GitHub only, `gfpop` and
`cpss` were removed from CRAN, deep-learning detectors live in Python, and
your in-house method lives with you.

That is a smaller problem than it looks, because almost nothing in this
package is about the detectors. `autoplot()`, the geoms, `tidy()`,
`glance()`, `augment()`, `cpt_metrics()`, `cpt_consensus()`,
`cpt_stability()`, `cpt_benchmark()` and `cpt_report()` all speak to the
`ggcpt` contract, not to any engine. Two functions let anything into it.

# `as_ggcpt()`: changepoints in, result out

The one-off case: you have locations from somewhere else and want the
package's machinery.

```{r as-ggcpt}
set.seed(2026)
x <- c(rnorm(100), rnorm(100, 4), rnorm(100, 1))

# Pretend these came from a Python detector, a paper, or an analyst.
external <- c(101, 199)

fit <- as_ggcpt(external, x, method = "ruptures::Pelt",
                cp_convention = "right")
fit
```

`cp_convention = "right"` matters: some engines report the first index of
the new segment and some the last index of the old one, and getting it
wrong shifts every location by one. The conversion happens on the way in,
so the stored result is always on this package's convention.

Everything now works:

```{r as-ggcpt-uses, fig.alt = "Series with the externally supplied changepoints drawn as vertical rules"}
tidy(fit)
glance(fit)
cpt_metrics(tidy(fit)$cp, truth = c(100, 200), n = 300)
autoplot(fit, show_segments = TRUE)
```

`as_ggcpt()` runs the same contract checks as every built-in wrapper —
sorting, de-duplication, range checks, aligned extra columns, derived
segments — so an external result cannot violate the invariants the rest of
the package relies on:

```{r as-ggcpt-contract}
tidy(as_ggcpt(c(150, 150, 9999, NA, 50), x))$cp
```

It also takes the optional extras, so a detector that reports uncertainty
does not have to throw it away:

```{r as-ggcpt-extras}
with_ci <- as_ggcpt(c(100, 200), x, method = "external",
                    ci = cbind(c(95, 192), c(107, 205)),
                    extra = list(score = c(12.4, 8.1)))
tidy(with_ci)
```

# `cpt_register_method()`: dispatch by name

The repeatable case: you want `cpt_detect(x, method = "yours")` and
everything that keys off a method name.

```{r register}
cpt_register_method(
  "biggest_jump",
  fn = function(x, window = 1, ...) {
    d <- abs(diff(as.numeric(x)))
    which.max(stats::filter(d, rep(1, window) / window, sides = 2))
  },
  change_in = "mean",
  engine = "example",
  citation = "No citation supplied (illustration only)."
)

res <- cpt_detect(x, method = "biggest_jump", window = 5)
res
```

Note what the print method says. A registered method is **visibly**
user-supplied everywhere it appears:

```{r register-visible}
subset(cpt_methods(), status == "registered")
cpt_cite("biggest_jump")
```

The package validates the *shape* of what your function returns; it does
not and cannot validate the method. That distinction is the reason for the
labelling, and it is why `cpt_cite()` says plainly when no citation was
given rather than inventing one.

Registered methods take part in everything that keys off a method name:

```{r register-tools}
cpt_consensus(x, methods = c("pelt", "binseg", "biggest_jump"),
              min_votes = 2)
cpt_benchmark(cpt_datasets(n = 200, seed = 1, names = c("step", "teeth")),
              methods = c("pelt", "biggest_jump"), progress = FALSE)
```

```{r unregister}
cpt_unregister_method("biggest_jump")
```

Registration is session state. It is deliberately not persisted to disk: a
script that behaved differently depending on what some earlier script had
run would be worse than the problem it solved.

## Returning a finished result

`fn` may return a bare vector of indices, as above, or a finished `ggcpt`
built with `as_ggcpt()` — which is how you pass along an engine's
confidence intervals, fitted signal or raw fit object:

```{r register-full}
cpt_register_method(
  "smoothed_jump",
  fn = function(x, ...) {
    sm <- stats::filter(x, rep(1, 11) / 11, sides = 2)
    sm[is.na(sm)] <- x[is.na(sm)]
    as_ggcpt(which.max(abs(diff(sm))), x, fitted = as.numeric(sm))
  },
  engine = "example"
)
autoplot(cpt_detect(x, method = "smoothed_jump"), show_fit = TRUE)
cpt_unregister_method("smoothed_jump")
```

# Recipe: a Python detector through reticulate

`ruptures` is the standard Python changepoint library. The recipe is four
lines, and this package never depends on `reticulate` to support it:

```{r reticulate, eval = FALSE}
library(reticulate)
rpt <- import("ruptures")

cpt_register_method(
  "ruptures_pelt",
  fn = function(x, model = "l2", pen = 10, ...) {
    algo <- rpt$Pelt(model = model)$fit(matrix(as.numeric(x), ncol = 1))
    # ruptures returns 1-based *right* endpoints, with n as the last entry
    as.integer(unlist(algo$predict(pen = pen)))
  },
  change_in = "mean",
  engine = "ruptures (Python)",
  cp_convention = "left",
  citation = paste("Truong, C., Oudre, L. and Vayatis, N. (2020).",
                   "Selective review of offline change point detection",
                   "methods. Signal Processing, 167, 107299.")
)

cpt_detect(x, method = "ruptures_pelt", pen = 20)
```

The same shape works for a detector loaded from GitHub, a compiled binary
called through `system2()`, or a neural network scored in `torch`: if it
returns changepoint indices, it joins the grammar.

# Reducing the install

The other side of extensibility is not needing everything. Only three
engines are hard dependencies. `cpt_methods()` reports what is installed,
and `cpt_install_engines()` installs a family at a time:

```{r install, eval = FALSE}
cpt_install_engines("bayesian")
cpt_install_engines(c("highdim", "functional"), dry_run = TRUE)
```

```{r install-status}
tab <- cpt_methods()
table(status = tab$status, installed = tab$installed, useNA = "ifany")
```

# What the contract is

If you are writing a wrapper of your own, this is the whole contract, and
`as_ggcpt()` enforces all of it:

- `$changepoints` is a tibble with at least `cp` (integer, sorted,
  de-duplicated, in `1..n-1`, on the "left" convention) and `cp_value`;
  engine extras such as `ci_lower`/`ci_upper` or `posterior_prob` are
  additional columns aligned to those rows.
- `$segments` has one row per segment (`seg_id`, `start`, `end`, `n`,
  `param_estimate`) and always has one more row than `$changepoints`.
- `$data` has `index` (positions `1..n`) and `value`, plus `fitted` when
  an engine supplies a signal and `index_value` when the result carries a
  time index.
- `$method`, `$change_in`, `$penalty`, `$cp_convention` are length-one
  metadata; `$fit` is the raw upstream object.
- The optional slots — `data_wide`, `index`, `regions`, `diagnostics`,
  `registered` — are absent unless something supplies them, so
  `is.null(fit$regions)` is the test for "this engine does not do
  regions".

# References
