---
title: "The ggchangepoint Feature Tour: A Complete Map of the Package Surface"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The ggchangepoint Feature Tour: A Complete Map of the Package Surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 72,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint feature tour plot"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_light())

# Optional engines live in Suggests; every chunk that needs one is guarded so
# the vignette builds on any installation.
has_fpop      <- requireNamespace("fpop", quietly = TRUE)
has_wbs       <- requireNamespace("wbs", quietly = TRUE)
has_breakfast <- requireNamespace("breakfast", quietly = TRUE)
has_not       <- requireNamespace("not", quietly = TRUE)
has_mosum     <- requireNamespace("mosum", quietly = TRUE)
has_idetect   <- requireNamespace("IDetect", quietly = TRUE)
has_stepR     <- requireNamespace("stepR", quietly = TRUE)
has_cpop      <- requireNamespace("cpop", quietly = TRUE)
has_bcp       <- requireNamespace("bcp", quietly = TRUE)
has_ocp       <- requireNamespace("ocp", quietly = TRUE)
# Rbeast (<= 1.0.2) can crash the vignette-building subprocess on Windows
# (its C internals have known memory-state issues); run it elsewhere only.
has_rbeast    <- requireNamespace("Rbeast", quietly = TRUE) &&
  .Platform$OS.type != "windows"
has_cpm       <- requireNamespace("cpm", quietly = TRUE)
has_kcprs     <- requireNamespace("kcpRS", quietly = TRUE)
has_cptnonpar <- requireNamespace("CptNonPar", quietly = TRUE)
has_decafs    <- requireNamespace("DeCAFS", quietly = TRUE)
has_snseg     <- requireNamespace("SNSeg", quietly = TRUE)
has_inspect   <- requireNamespace("InspectChangepoint", quietly = TRUE)
has_geomcp    <- requireNamespace("changepoint.geo", quietly = TRUE)
has_struc     <- requireNamespace("strucchange", quietly = TRUE)
has_segmented <- requireNamespace("segmented", quietly = TRUE)
has_envcpt    <- requireNamespace("EnvCpt", quietly = TRUE)
has_fastcpd   <- requireNamespace("fastcpd", quietly = TRUE)
has_plotly    <- requireNamespace("plotly", quietly = TRUE)
```

# Abstract

**ggchangepoint** provides a unified, tidy, `ggplot2`-native interface to
changepoint detection in R: one dispatcher (`cpt_detect()`) covering 50
methods across six methodological families, one result class (`ggcpt`) with
a stable tidy contract, and one visualisation entry point (`autoplot()`)
that draws everything a method reports — including confidence intervals and
posterior probabilities [@wickham2016ggplot2; @robinson2017broom]. This
vignette is the *feature tour*: it visits **every exported function** in the
package at the point where it belongs in the workflow, so a reader can map
the full surface in one sitting.

Six companion vignettes go deeper: `vignette("introduction")` develops the
methodology, `vignette("comparison")` treats method comparison and
evaluation, `vignette("inference")` covers confidence intervals,
significance regions, selecting the number of changepoints and the
diagnostic displays, `vignette("supervised")` covers labelled regions and
learned penalties, `vignette("extending")` shows how to bring an external
detector into the same grammar, and `vignette("monitoring")` covers
sequential detection on a stream and its detection-delay accounting.

> **New in 0.5.0.** Nineteen further engines; a time index that survives
> the round trip (`cpt_detect(x, index = dates)` and `ts`/`xts`/`zoo`/
> `tsibble` input); significance regions and the `cpt_confint()` /
> `cpt_test()` inference pair; `cpt_select()` for the number of
> changepoints; influence, sensitivity, statistic, solution-path and
> scale-space diagnostics; supervised detection; consensus and method
> recommendation; event annotation and reports; benchmarking; sequential
> monitoring; power analysis; and `as_ggcpt()` / `cpt_register_method()`
> for detectors this package does not wrap. Each has its own section
> below or its own vignette.

```{r data}
set.seed(2026)
x <- c(rnorm(100, 0, 1), rnorm(100, 6, 1))    # one mean shift at t = 100
x3 <- c(rnorm(100), rnorm(100, 4), rnorm(100, 1))  # shifts at 100, 200
```

# The result object and its methods

Every detector returns a `ggcpt` object: a list carrying the tidy
`changepoints` tibble (`cp` = last index of the left segment, `cp_value` =
series value there, plus any method-specific columns), a `segments` table,
the `data`, the raw engine `fit`, and metadata (method, `change_in`,
penalty, convention, runtime). `new_ggcpt()` is the low-level constructor
and `is_ggcpt()` the class test; most users never call either directly.

```{r class}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
is_ggcpt(res)
print(res)
```

```{r constructor}
manual <- new_ggcpt(
  changepoints = tibble::tibble(cp = 100L, cp_value = x[100]),
  data = tibble::tibble(index = seq_along(x), value = x),
  method = "manual"
)
is_ggcpt(manual)
```

The `broom` verbs give one row per changepoint (`tidy()`), a one-row model
summary (`glance()`), and the data augmented with segment ids, fitted
levels, residuals, and a changepoint flag (`augment()`):

```{r broom}
tidy(res)
glance(res)
head(augment(res))
```

The remaining S3 surface: a human-readable digest, tibble/data-frame
coercion, a one-line format, and a base-`plot()` fallback that delegates to
`autoplot()`.

```{r s3}
summary(res)
as_tibble(res)
head(as.data.frame(res))
format(res)
```

```{r plot-fallback, fig.alt = "The default plot: the series with its changepoint rules, drawn by plot() delegating to autoplot()"}
plot(res)
```

# Unified detection

`cpt_detect()` is the recommended entry point: pick a `method`, say what the
change is in (`change_in` = `"mean"`, `"var"`, `"meanvar"`, `"slope"`,
`"distribution"`, `"covariance"`, `"network"`, `"regression"` or
`"seasonality"` — the values `cpt_methods()` lists in its `supports`
column), and optionally set a `penalty`. Incompatible
`method`/`change_in` combinations error — they are never silently
substituted.

```{r detect}
cpt_detect(x, method = "binseg", change_in = "mean")
```

Anything else in `...` reaches the underlying wrapper, and takes precedence
over the value the dispatcher would otherwise derive from `change_in` —
here the NOT contrast is set directly rather than inherited:

```{r detect-dots, eval = has_not}
tidy(cpt_detect(x, method = "not", contrast = "pcwsLinMean"))
```

`cpt_methods()` is the live capability table: every method the package
knows, its engine package, what it can detect, and whether the engine is
installed. Five rows carry status `"planned"` rather than `"available"`:
their engines (`gfpop`, `robseg`, `FOCuS`, `hdbinseg`, `changeforest`) are
not on CRAN, so they are documented as future work and are *not* wired to
`cpt_detect()`.

```{r methods}
print(cpt_methods(), n = Inf)
```

`cpt_penalty()` constructs standard penalty values for the engines that
take numeric penalties; see its help for the per-engine penalty semantics.

```{r penalty}
cpt_penalty("BIC", n = 200)
cpt_penalty("MBIC", n = 200)
cpt_penalty("Manual", value = 10)
```

# Engine wrappers

Each engine also has a direct wrapper exposing its native arguments. Every
wrapper from 0.2.0 onwards returns a `ggcpt` object; only the two original
wrappers below predate the class and still return a bare tibble.

## The classical core (0.1.0)

`cpt_wrapper()` and `ecp_wrapper()` are the original interface to the
*changepoint*/*changepoint.np* [@killick2012pelt; @killick2014changepoint;
@haynes2017computationally] and *ecp* [@matteson2014nonparametric;
@james2014ecp] engines; they return bare tibbles for backward
compatibility.

```{r classic}
cpt_wrapper(x, change_in = "mean", cp_method = "PELT")
ecp_wrapper(x, algorithm = "divisive", seed = 1)
```

## The search and pruning wave (0.2.0)

Seven multiscale, randomised, and functional-pruning engines
[@fryzlewicz2014wild; @fryzlewicz2020detecting; @baranowski2019narrowest;
@eichinger2018mosum; @anastasiou2022idetect; @fryzlewicz2018tail;
@maidstone2017optimal], one call each:

```{r fpop, eval = has_fpop}
fpop_wrapper(x, penalty = 2 * log(length(x)))
```

```{r wbs, eval = has_wbs}
wbs_wrapper(x, n_intervals = 2000, seed = 1)
```

```{r wbs2, eval = has_breakfast}
wbs2_wrapper(x)
```

```{r not, eval = has_not}
not_wrapper(x, contrast = "pcwsConstMean", seed = 1)
```

```{r mosum, eval = has_mosum}
mosum_wrapper(x)
mosum_wrapper(x3, multiscale = TRUE)
```

```{r idetect, eval = has_idetect}
idetect_wrapper(x, seed = 1)
```

```{r tguh, eval = has_breakfast}
tguh_wrapper(x)
```

## The engine wave (0.4.0)

**Multiscale inference with confidence.** `smuce_wrapper()` implements SMUCE
[@frick2014smuce], which bounds the probability of over-estimating the number
of changes at the level `alpha` and returns a confidence interval for every
changepoint location (`ci_lower`/`ci_upper`); `family = "hsmuce"` switches to
the heteroskedastic extension HSMUCE [@pein2017hsmuce], which estimates a
variance per segment:

```{r smuce, eval = has_stepR}
res_smuce <- smuce_wrapper(x, alpha = 0.5)
tidy(res_smuce)
```

**Change in slope.** `cpop_wrapper()` performs exact penalised estimation of
a continuous piecewise-linear mean [@fearnhead2019cpop; @fearnhead2024cpop]:

```{r cpop, eval = has_cpop}
y_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200)
res_cpop <- cpop_wrapper(y_slope)
tidy(res_cpop)
```

```{r cpop-fallback, include = FALSE, eval = !has_cpop}
# y_slope is reused by the segmented example below
y_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200)
```

**Bayesian detection.** `bcp_wrapper()` implements the Barry–Hartigan
product-partition model [@barry1993bayesian; @erdman2007bcp];
`bocpd_wrapper()` runs Bayesian online changepoint detection over the
run-length posterior [@adams2007bocpd]; `beast_wrapper()` wraps the BEAST
Bayesian model-averaging ensemble [@zhao2019beast]. The first and third keep
the locations whose posterior probability clears `prob_threshold` and record
it in a `posterior_prob` column; BOCPD returns the maximum a posteriori
changepoint set.

```{r bcp, eval = has_bcp}
res_bcp <- bcp_wrapper(x, seed = 1)
tidy(res_bcp)
```

```{r bocpd, eval = has_ocp}
res_bocpd <- bocpd_wrapper(x)
tidy(res_bocpd)
```

```{r beast, eval = has_rbeast}
res_beast <- beast_wrapper(x, seed = 1)
tidy(res_beast)
```

**Sequential and kernel nonparametrics.** `cpm_wrapper()` runs
distribution-free sequential tests and reports, alongside each estimated
location, the `detection_time` at which a stream monitor would have flagged
it [@ross2015cpm]; `kcp_wrapper()` applies kernel change point analysis to
running statistics (mean, variance, autocorrelation, correlation)
[@arlot2019kernel; @cabrieto2018kcprs]; `npmojo_wrapper()` detects
distributional changes under serial dependence [@mcgonigle2023npmojo]:

```{r cpm, eval = has_cpm}
tidy(cpm_wrapper(x, cpm_type = "Mann-Whitney"))
```

```{r kcp, eval = has_kcprs}
tidy(kcp_wrapper(x, running_stat = "mean", nperm = 100, seed = 1))
```

```{r npmojo, eval = has_cptnonpar}
tidy(npmojo_wrapper(x))
```

**Robustness to drift and dependence.** `decafs_wrapper()` detects abrupt
changes when the signal also drifts and the noise is autocorrelated
[@romano2022decafs]; `sn_wrapper()` segments a chosen `parameter` (mean,
variance, autocorrelation, or bivariate correlation) by self-normalisation,
avoiding long-run variance estimation entirely [@zhao2022snseg];
`envcpt_wrapper()` reports changepoints only when a changepoint model beats
the trend and autoregressive alternatives [@beaulieu2018envcpt]:

```{r decafs, eval = has_decafs}
tidy(decafs_wrapper(x))
```

```{r sn, eval = has_snseg}
tidy(sn_wrapper(x3, parameter = "mean"))
```

```{r envcpt, eval = has_envcpt}
res_env <- envcpt_wrapper(x, models = c("mean", "meancpt", "trendcpt"))
tidy(res_env)
res_env$penalty$type   # criterion and winning model
```

**High-dimensional and multivariate.** These wrappers take a matrix or data
frame with one row per time point. `inspect_wrapper()` finds sparse mean
changes by projection [@wang2018inspect]; `ocd_wrapper()` monitors a
high-dimensional stream online [@chen2022ocd]; `geomcp_wrapper()` maps each
observation to a distance and an angle and segments both
[@grundy2020geomcp]; multivariate `ecp` input flows through `cpt_detect()`
unchanged. The univariate wrappers, by contrast, reject a multi-column
argument rather than silently flattening it.

```{r hd-data}
set.seed(1)
X <- cbind(a = c(rnorm(80), rnorm(80, 4)),
           b = c(rnorm(80), rnorm(80, -3)),
           c = rnorm(160))
```

```{r inspect, eval = has_inspect}
res_hd <- inspect_wrapper(X)
tidy(res_hd)
```

```{r ocd, eval = FALSE}
# Online detection: the reported locations are declaration times (the change
# plus the detection delay), in a `declared_at` column. Monte Carlo threshold
# calibration makes this the slowest wrapper, so it is shown but not run
# here. It needs at least two coordinates and rejects a univariate series.
res_ocd <- ocd_wrapper(X, mc_reps = 100)
tidy(res_ocd)
```

```{r geomcp, eval = has_geomcp}
tidy(geomcp_wrapper(X))
```

```{r ecp-mv}
tidy(cpt_detect(X, method = "ecp", seed = 1))
```

**Regression structure.** `strucchange_wrapper()` dates Bai–Perron breaks,
either in a bare series or in the coefficients of a formula
[@bai1998estimating; @bai2003computation; @zeileis2002strucchange];
`segmented_wrapper()` fits a continuous broken-line regression, so the
change it reports is a kink in the trend rather than a jump in the level
[@muggeo2003segmented; @muggeo2008segmented]. Both carry a confidence
interval for every break:

```{r strucchange, eval = has_struc}
tidy(strucchange_wrapper(x))
```

```{r segmented, eval = has_segmented}
tidy(segmented_wrapper(y_slope, npsi = 1, seed = 1))
```

**The modern PELT family.** `fastcpd_wrapper()` exposes the *fastcpd*
engine, whose `family` argument covers changes in the mean, the variance, or
both, as well as changes in a fitted AR/ARMA/GARCH model [@li2024fastcpd]:

```{r fastcpd, eval = has_fastcpd}
tidy(fastcpd_wrapper(x, family = "mean"))
```

# The penalty path

Rather than guessing one penalty, `cpt_crops()` computes *every* optimal
segmentation over a penalty interval (the CROPS algorithm, via
*changepoint* [@killick2014changepoint]) and returns a `ggcpt_path` object
with its own `print()`, `tidy()`, and three `autoplot()` types:

```{r crops}
path <- cpt_crops(x3)
path
tidy(path)
```

```{r crops-elbow, fig.alt = "CROPS cost elbow: segmentation cost against the number of changepoints for each solution on the penalty path"}
autoplot(path)                          # the cost elbow
```

```{r crops-path, fig.alt = "Changepoint locations against the penalty, showing which locations survive as the penalty rises"}
autoplot(path, type = "path")           # changepoints vs penalty
```

```{r crops-seg, fig.alt = "The candidate CROPS segmentations, faceted by number of changepoints, each panel showing the series with that solution's changepoints"}
autoplot(path, type = "segmentations")  # the candidate models themselves
```

# The visualisation layer

`autoplot()` renders any `ggcpt`. Options: `show_segments` (fitted segment
means), `show_fit` (the engine's own fitted signal, where provided — the nine
engines `cpt_methods()` marks in its `fitted` column: SMUCE, HSMUCE, CPOP,
bcp, BEAST, DeCAFS, segmented, mcp, bfast), `show_ci` (changepoint-location
confidence intervals, where provided — the seven marked in its `ci` column:
SMUCE, HSMUCE, strucchange, segmented, mcp, bfast, taylor; `nsp` is marked
there too but reports a region, drawn by `show_regions`),
`show_points`/`show_line`, an `index` for a date axis, and the
`cptline_*` styling arguments (`cptline_color`, `cptline_alpha`,
`cptline_type`, `cptline_linewidth`). Asking for an overlay the result
cannot supply warns rather than failing silently.

```{r autoplot, fig.alt = "Series with its changepoints drawn as firebrick rules and the fitted segment levels as horizontal steps"}
autoplot(res, show_segments = TRUE, cptline_color = "firebrick",
         cptline_type = "dashed", cptline_linewidth = 0.8)
```

```{r autoplot-ci, eval = has_stepR, fig.alt = "SMUCE step fit overlaid on the series, with horizontal whiskers giving each changepoint's confidence interval"}
autoplot(res_smuce, show_ci = TRUE, show_fit = TRUE)
```

Multivariate results facet automatically:

```{r autoplot-mv, eval = has_inspect, fig.height = 6, fig.alt = "Faceted small-multiples, one panel per coordinate, sharing the detected changepoint rules"}
autoplot(res_hd)
```

The composable layers work inside any ggplot pipeline:
`geom_changepoint()` draws vertical rules, `geom_cpt_segment()` draws
segment levels, `geom_cpt_ci()` draws horizontal interval whiskers, and
`stat_changepoint()` runs detection *inside* the plot. `theme_ggcpt()` is a
publication theme and `annotate_segments()` shades alternating segments.

```{r geoms, fig.alt = "The same plot assembled layer by layer: shaded segment bands behind the series, a red rule at each changepoint, and the fitted level drawn as a segment across each span"}
cp_tbl <- tidy(res)
df <- data.frame(index = seq_along(x), value = x)

ggplot(df, aes(index, value)) +
  annotate_segments(cp = cp_tbl$cp, n = length(x)) +
  geom_line() +
  geom_changepoint(data = cp_tbl, aes(xintercept = cp), color = "red") +
  geom_cpt_segment(
    data = res$segments,
    aes(x = start, xend = end, y = param_estimate, yend = param_estimate),
    inherit.aes = FALSE, color = "darkred", linewidth = 1
  ) +
  theme_ggcpt()
```

```{r stat, fig.alt = "Series with changepoints added by stat_changepoint(), which runs the detector inside the layer"}
ggplot(df, aes(index, value)) +
  geom_line() +
  stat_changepoint(method = "pelt", color = "blue")
```

```{r geom-ci, eval = has_stepR, fig.alt = "Series in grey with a row of blue horizontal whiskers below it, one per changepoint, each spanning that changepoint's confidence interval"}
ci_tbl <- tidy(res_smuce)
ci_tbl$y_pos <- min(x) - 1
ggplot(df, aes(index, value)) +
  geom_line(color = "grey60") +
  geom_cpt_ci(
    data = ci_tbl,
    aes(y = y_pos, xmin = ci_lower, xmax = ci_upper),
    width = 0.6, color = "blue", inherit.aes = FALSE
  ) +
  geom_changepoint(data = ci_tbl, aes(xintercept = cp), color = "blue")
```

`ggcptplot()` and `ggecpplot()` are the original one-call plots for the two
classical engines:

```{r original-plots, fig.alt = "The original one-call plot: the series with its PELT changepoints marked, drawn by ggcptplot()"}
ggcptplot(x, change_in = "mean", cp_method = "PELT")
```

```{r ggecpplot, fig.alt = "The original one-call ecp plot: the series with the divisive algorithm's changepoints marked"}
ggecpplot(x, algorithm = "divisive", seed = 1)
```

The Bayesian engines get the field's signature displays:
`ggcpt_posterior()` shows the posterior mean over the series and the
per-location changepoint probability (for `bcp` and BEAST results);
`ggcpt_runlength()` shows the BOCPD run-length posterior as a heatmap.

```{r posterior, eval = has_bcp, fig.height = 6, fig.alt = "Two-panel Bayesian display: the series with its posterior mean above, per-location posterior changepoint probability below"}
ggcpt_posterior(res_bcp)
```

```{r runlength, eval = has_ocp, fig.alt = "Run-length heatmap: posterior probability of each run length over time"}
ggcpt_runlength(res_bocpd)
```

Finally, `ggcpt_interactive()` turns any result (or any ggplot built from
one) into a `plotly` widget with values on hover. The widget itself is not
embedded here, only its class:

```{r interactive, eval = has_plotly}
class(ggcpt_interactive(res))   # requires plotly
```

# Comparing methods

`ggcpt_compare()` runs several detectors on the same series and renders them
faceted (default) or overlaid, honouring `future::plan()` for parallel
execution; `ggcpt_compare_table()` returns the tidy union of the same runs.
A method that finds nothing keeps its panel and contributes an `NA` row —
"no changepoints" is a result, not a missing one.

```{r compare, fig.height = 6, fig.alt = "One panel per method, each showing the same series with that method's changepoints"}
ggcpt_compare(x, methods = c("pelt", "binseg", "amoc"))
```

```{r compare-overlay, fig.alt = "All three methods' changepoints overlaid on one copy of the series, coloured by method"}
ggcpt_compare(x, methods = c("pelt", "binseg"), layout = "overlay")
```

```{r compare-table}
ggcpt_compare_table(x, methods = c("pelt", "binseg", "amoc"))
```

# Batch detection and stability

`cpt_batch()` runs one detector over many series (matrix, data frame, or
list of vectors), also under `future::plan()`, and returns a tidy
`ggcpt_batch` tibble with list-columns for the per-series changepoints and
the `ggcpt` objects themselves:

```{r batch, fig.alt = "Small-multiples of a panel of series, each with its own detected changepoints"}
XB <- cbind(shifted = x, pure_noise = rnorm(200))
batch <- cpt_batch(XB, method = "pelt")
batch
tidy(batch)
autoplot(batch)
```

`cpt_stability()` quantifies how fragile a segmentation is: it resamples
residuals within the fitted segments, re-runs the detector, and reports the
re-detection frequency at every location — a model-agnostic confidence
signal for engines with no native intervals:

```{r stability, fig.alt = "Bootstrap detection-frequency profile along the series, with the original changepoints marked as dashed rules"}
st <- cpt_stability(x, method = "pelt", B = 30, seed = 1)
st
autoplot(st)
```

# Evaluation against ground truth

`cpt_metrics()` scores predictions against a known truth: precision, recall
and F1 under one-to-one matching, the covering metric and adjusted Rand
index in the conventions of @van2020evaluation, Hausdorff distance,
annotation error, and matched MAE/RMSE. `cpt_metrics_annotated()` averages
over multiple annotators, and `ggcpt_eval()` draws the agreement (true
positives, false positives, and misses):

```{r metrics}
truth <- 100
pred <- tidy(res)$cp
cpt_metrics(pred, truth, n = length(x), margin = 5)
cpt_metrics_annotated(pred, list(100, 101, 99), n = length(x))
```

```{r eval-plot, fig.alt = "Series with predictions coloured as true positives and false positives, shaded tolerance windows around each true changepoint, and missed truths as dashed rules"}
ggcpt_eval(pred, truth, x, margin = 5)
```

# Simulation and canonical signals

`cpt_simulate()` (alias `rcpt()`) generates series with known changepoints
in mean, variance, both, or slope, under Gaussian, Student-t, AR(1), or
random-walk noise; the truth travels with the data in the
`true_changepoints` and `true_segments` attributes. Five canonical test
signals from the literature ship ready-made: `signal_blocks()` (the
Donoho–Johnstone blocks signal [@donoho1994ideal]), `signal_fms()`,
`signal_mix()`, `signal_teeth()`, and `signal_stairs()`.

```{r simulate}
sim <- cpt_simulate(300, changepoints = c(100, 200), change_in = "mean",
                    params = c(0, 5, 1), seed = 1)
attr(sim, "true_changepoints")
sim2 <- rcpt(300, changepoints = 150, params = c(0, 3), seed = 2)  # the alias
attr(sim2, "true_changepoints")

signals <- list(blocks = signal_blocks(1024, seed = 1),
                fms    = signal_fms(500, seed = 1),
                mix    = signal_mix(500, seed = 1),
                teeth  = signal_teeth(400, seed = 1),
                stairs = signal_stairs(500, seed = 1))
vapply(signals, function(s) length(attr(s, "true_changepoints")), integer(1))
```

```{r blocks-plot, fig.alt = "The Donoho-Johnstone blocks test signal, with its true changepoints marked by thin blue rules"}
blocks <- signals$blocks
ggplot(blocks, aes(index, value)) +
  geom_line(color = "grey50") +
  geom_vline(xintercept = attr(blocks, "true_changepoints"),
             color = "blue", linewidth = 0.3) +
  labs(title = "The blocks test signal with its true changepoints")
```

# Citing the methodology

`cpt_cite()` returns the verified reference(s) behind a result or a method
name, so a write-up can cite the right paper without leaving R; called with
no argument it returns the whole method-to-reference table.

```{r cite}
cpt_cite("pelt")
cpt_cite(res)
```

# The 0.5.0 surface

The sections above are the 0.4.0 tour. What follows visits the rest of the
exported surface, one example each; four companion vignettes develop these
properly.

## Time indices

`cpt_detect(index = )` attaches a clock. Detection still runs on positions,
because every engine assumes an equally spaced sequence, but `tidy()`,
`augment()`, `autoplot()` and the reporting functions all speak in the
user's units. `as_cpt_series()` is the coercion underneath, and it accepts
`ts`, `xts`, `zoo` and `tsibble` directly.

```{r index-tour}
dates <- as.Date("2020-01-01") + seq_along(x) - 1
dated <- cpt_detect(x, method = "pelt", index = dates)
tidy(dated)
cpt_detect(data.frame(day = dates, v = x), y = v, index = day,
           method = "pelt")$changepoints$cp_index
```

`as_cpt_series()` is that coercion on its own, returning the values, the
index and a label for the axis:

```{r as-cpt-series}
str(as_cpt_series(x, index = dates), max.level = 1)
as_cpt_series(ts(x, start = c(2020, 1), frequency = 12))$index_label
```

## Inference

`nsp_wrapper()` returns significance regions, read with `cpt_regions()` and
drawn by `geom_cpt_region()`. `cpt_confint()` unifies four interval routes;
`cpt_test()` attaches a test and flags whether it is selection-adjusted.

```{r inference-tour}
cpt_confint(res, method = "bootstrap", B = 30, seed = 1)
suppressWarnings(cpt_test(res))[, c("cp", "estimate", "p_value",
                                    "selection_adjusted")]
```

```{r nsp-tour, eval = requireNamespace("nsp", quietly = TRUE), fig.alt = "Series with a shaded band marking an NSP significance region"}
autoplot(nsp_wrapper(x, M = 100, seed = 1))
```

## Selecting the number of changepoints

```{r select-tour, fig.alt = "Criterion value against the number of changepoints, with the chosen model marked"}
sel <- cpt_select(x3, criterion = "bic", k_max = 6)
sel$k
autoplot(sel)
```

## Diagnostics

`cpt_influence()` and `cpt_leverage()` say which observation drives the
answer; `cpt_sensitivity()` says which setting does; `cpt_statistic()`,
`cpt_solution_path()` and `cpt_scale_space()` (with their `ggcpt_*()`
plots, reachable from `autoplot(type = )`) say what the detector actually
computed.

```{r diag-tour}
inf <- cpt_influence(res, engine = "recompute", subset = seq(1, 200, by = 20))
head(cpt_leverage(inf), 3)
cpt_sensitivity(x, method = "pelt", over = list(penalty = c(4, 40)))$grid[, 1:2]
```

```{r stat-tour, eval = has_mosum, fig.alt = "Two-panel display: the series above, the MOSUM statistic below"}
autoplot(cpt_detect(x, method = "mosum"), type = "statistic")
```

## Supervised detection

```{r supervised-tour}
labs <- cpt_labels(c(80, 140), c(120, 190), c("one_change", "no_change"))
cpt_label_error(res, labs)[, c("start", "end", "change", "status")]
```

`as_cpt_labels()` converts a plain ground-truth set into labels,
`cpt_label_error_curve()` traces the errors across a penalty grid, and
`cpt_learn_penalty()` fits a penalty from several labelled series — see
`vignette("supervised", package = "ggchangepoint")`.

## Choosing and combining methods

```{r consensus-tour}
cpt_recommend(noise = "heavy")
tidy(cpt_consensus(x3, methods = c("pelt", "binseg", "amoc")))
```

## Communication

```{r events-tour}
cpt_annotate_events(dated,
                    data.frame(when = as.Date("2020-04-10"),
                               what = "policy change"))
```

`cpt_report()` assembles the whole analysis into markdown, and `cpt_gt()`
renders a publication table (falling back to a tibble, with a note, when
`gt` is not installed):

```{r gt-tour, eval = requireNamespace("gt", quietly = TRUE)}
cpt_gt(res, title = "Detected changepoints")
```

`geom_cpt_event()` draws the events. `scale_colour_cpt()`,
`scale_fill_cpt()`, `scale_linetype_cpt()`, `scale_fill_cpt_label()` and
`scale_colour_cpt_label()` are the colour-vision-safe scales — the
Okabe-Ito palette, plus a linetype scale so that colour is never the only
channel carrying a distinction. All three at once, on the fitted
segmentation:

```{r accessibility-tour, fig.alt = "Fitted segment levels drawn with the Okabe-Ito palette, a matching fill band and a redundant linetype per segment"}
aug <- augment(res)
ggplot(aug, aes(index, .fitted, colour = factor(seg_id),
                linetype = factor(seg_id))) +
  geom_line(aes(y = value), colour = "grey75", linetype = "solid") +
  geom_ribbon(aes(ymin = .fitted - 1, ymax = .fitted + 1,
                  fill = factor(seg_id)), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_cpt() + scale_fill_cpt() + scale_linetype_cpt() +
  labs(x = "Index", y = "Value", colour = "Segment", fill = "Segment",
       linetype = "Segment")
```

`theme_ggcpt()` is unchanged.

## Benchmarking

```{r benchmark-tour, fig.alt = "Heatmap of the covering metric for each method on each benchmark dataset"}
bm <- cpt_benchmark(cpt_datasets(n = 200, seed = 1,
                                 names = c("step", "teeth", "ar1")),
                    methods = c("pelt", "binseg", "amoc"), progress = FALSE)
autoplot(bm)
cpt_annotations(cpt_datasets(n = 200, names = "step"))
```

`cpt_load_tcpd()` downloads and caches the Turing Change Point Dataset; it
is not run here because it needs a network connection.

## Streaming

```{r monitor-tour}
mon <- cpt_replay(c(rnorm(200), rnorm(200, 3)), method = "edetector")
alarms(mon)
cpt_delay(mon, truth = 200)
```

`cpt_monitor()` and `cpt_update()` are the incremental form:
`mon <- cpt_update(mon, new_obs)` as data arrive.

## Power and study design

```{r power-tour, fig.alt = "Power curve: detection probability against change size"}
pw <- cpt_power(n = 200, jump = c(0.5, 1, 2), n_sim = 30, seed = 1)
autoplot(pw)
```

`cpt_min_detectable()` inverts the curve — it bisects for the change size
that reaches a target power, which is the number a pre-registration needs:

```{r min-detectable-tour}
cpt_min_detectable(n = 150, n_sim = 15, max_iter = 3, seed = 1)$jump
```

`cpt_scenarios()` builds the grid a simulation study varies over as data
rather than as nested loops, so it can be inspected, filtered and passed
straight to `cpt_benchmark()`:

```{r scenarios-tour}
cpt_scenarios(n = c(200, 500), jump = c(1, 2), as_datasets = FALSE)
```

## Extending

```{r extend-tour}
fit_ext <- as_ggcpt(c(100), x, method = "external_detector")
glance(fit_ext)[, c("method", "n_changepoints")]

cpt_register_method("biggest_jump",
                    fn = function(x, ...) which.max(abs(diff(x))),
                    engine = "example")
cpt_detect(x, method = "biggest_jump")$changepoints$cp
subset(cpt_methods(), status == "registered")[, c("method", "engine")]
cpt_registered_methods()
cpt_unregister_method("biggest_jump")
```

`cpt_install_engines()` installs a family of engines at a time:

```{r install-tour}
cpt_install_engines("bayesian", dry_run = TRUE)[, c("package",
                                                    "installed_before")]
```

# Closing note

This tour visited every exported function in the package. For the
framework's design, the mathematics of the wrapped methods, and worked
analyses, see `vignette("introduction", package = "ggchangepoint")`; for
method comparison and accuracy evaluation in depth, see
`vignette("comparison", package = "ggchangepoint")`; and for the 0.5.0
additions in depth, see `vignette("inference")`,
`vignette("supervised")`, `vignette("extending")` and
`vignette("monitoring")`.

# References
