---
title: "The ggchangepoint Feature Tour: A Complete Map of the Package Surface"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The ggchangepoint Feature Tour: A Complete Map of the Package Surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint feature tour plot"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_light())

# Optional engines live in Suggests; every chunk that needs one is guarded so
# the vignette builds on any installation.
has_fpop      <- requireNamespace("fpop", quietly = TRUE)
has_wbs       <- requireNamespace("wbs", quietly = TRUE)
has_breakfast <- requireNamespace("breakfast", quietly = TRUE)
has_not       <- requireNamespace("not", quietly = TRUE)
has_mosum     <- requireNamespace("mosum", quietly = TRUE)
has_idetect   <- requireNamespace("IDetect", quietly = TRUE)
has_stepR     <- requireNamespace("stepR", quietly = TRUE)
has_cpop      <- requireNamespace("cpop", quietly = TRUE)
has_bcp       <- requireNamespace("bcp", quietly = TRUE)
has_ocp       <- requireNamespace("ocp", quietly = TRUE)
# Rbeast (<= 1.0.2) can crash the vignette-building subprocess on Windows
# (its C internals have known memory-state issues); run it elsewhere only.
has_rbeast    <- requireNamespace("Rbeast", quietly = TRUE) &&
  .Platform$OS.type != "windows"
has_cpm       <- requireNamespace("cpm", quietly = TRUE)
has_kcprs     <- requireNamespace("kcpRS", quietly = TRUE)
has_cptnonpar <- requireNamespace("CptNonPar", quietly = TRUE)
has_decafs    <- requireNamespace("DeCAFS", quietly = TRUE)
has_snseg     <- requireNamespace("SNSeg", quietly = TRUE)
has_inspect   <- requireNamespace("InspectChangepoint", quietly = TRUE)
has_geomcp    <- requireNamespace("changepoint.geo", quietly = TRUE)
has_struc     <- requireNamespace("strucchange", quietly = TRUE)
has_segmented <- requireNamespace("segmented", quietly = TRUE)
has_envcpt    <- requireNamespace("EnvCpt", quietly = TRUE)
has_fastcpd   <- requireNamespace("fastcpd", quietly = TRUE)
has_plotly    <- requireNamespace("plotly", quietly = TRUE)
```

# Abstract

**ggchangepoint** provides a unified, tidy, `ggplot2`-native interface to
changepoint detection in R: one dispatcher (`cpt_detect()`) covering 31
methods across five methodological families, one result class (`ggcpt`) with
a stable tidy contract, and one visualization entry point (`autoplot()`)
that can draw everything a method reports — including confidence intervals
and posterior probabilities [@wickham2016ggplot2; @robinson2017broom]. This
vignette is the *feature tour*: it visits **every exported function** in the
package exactly where it belongs in the workflow, so that a reader can map
the full surface in one sitting. The companion vignettes develop the
methodology in depth (`vignette("introduction")`) and treat method
comparison and evaluation (`vignette("comparison")`).

```{r data}
set.seed(2026)
x <- c(rnorm(100, 0, 1), rnorm(100, 6, 1))    # one mean shift at t = 100
x3 <- c(rnorm(100), rnorm(100, 4), rnorm(100, 1))  # shifts at 100, 200
```

# The result object and its methods

Every detector returns a `ggcpt` object: a list carrying the tidy
`changepoints` tibble (`cp` = last index of the left segment, `cp_value` =
series value there, plus any method-specific columns), a `segments` table,
the `data`, and metadata (method, `change_in`, penalty, convention,
runtime). `new_ggcpt()` is the low-level constructor and `is_ggcpt()` the
class test; most users never call either directly.

```{r class}
res <- cpt_detect(x, method = "pelt", change_in = "mean")
is_ggcpt(res)
print(res)
```

```{r constructor}
manual <- new_ggcpt(
  changepoints = tibble::tibble(cp = 100L, cp_value = x[100]),
  data = tibble::tibble(index = seq_along(x), value = x),
  method = "manual"
)
is_ggcpt(manual)
```

The `broom` verbs give one row per changepoint (`tidy()`), a one-row model
summary (`glance()`), and the data augmented with segment ids, fitted
levels, and residuals (`augment()`):

```{r broom}
tidy(res)
glance(res)
head(augment(res))
```

The remaining S3 surface: a human-readable digest, tibble/data-frame
coercion, a one-line format, and a base-`plot()` fallback that delegates to
`autoplot()`.

```{r s3}
summary(res)
as_tibble(res)
head(as.data.frame(res))
format(res)
```

```{r plot-fallback}
plot(res)
```

# Unified detection

`cpt_detect()` is the recommended entry point: pick a `method`, say what the
change is in (`change_in` = `"mean"`, `"var"`, `"meanvar"`, `"slope"`, or
`"distribution"`), and optionally set a `penalty`. Incompatible
`method`/`change_in` combinations error — they are never silently
substituted.

```{r detect}
cpt_detect(x, method = "binseg", change_in = "mean")
```

`cpt_methods()` is the live capability table: every method the package
knows, its engine package, what it can detect, and whether the engine is
installed.

```{r methods}
cpt_methods()
```

`cpt_penalty()` constructs standard penalty values for the engines that
take numeric penalties; see its help for the per-engine penalty semantics.

```{r penalty}
cpt_penalty("BIC", n = 200)
cpt_penalty("MBIC", n = 200)
cpt_penalty("Manual", value = 10)
```

# Engine wrappers

Each engine also has a direct wrapper exposing its native arguments. All
return the same `ggcpt` object.

## The classical wave

`cpt_wrapper()` and `ecp_wrapper()` are the original 0.1.0 interface to
the *changepoint*/*changepoint.np* [@killick2012pelt;
@killick2014changepoint; @haynes2017computationally] and *ecp*
[@matteson2014nonparametric; @james2014ecp] engines; they return bare
tibbles for backward compatibility.

```{r classic}
cpt_wrapper(x, change_in = "mean", cp_method = "PELT")
ecp_wrapper(x, algorithm = "divisive", seed = 1)
```

The search and pruning engines added in 0.2.0
[@fryzlewicz2014wild; @baranowski2019narrowest; @eichinger2018mosum;
@anastasiou2022idetect; @fryzlewicz2022tail; @maidstone2017optimal],
one call each:

```{r fpop, eval = has_fpop}
fpop_wrapper(x, penalty = 2 * log(length(x)))
```

```{r wbs, eval = has_wbs}
wbs_wrapper(x, n_intervals = 2000, seed = 1)
```

```{r wbs2, eval = has_breakfast}
wbs2_wrapper(x)
```

```{r not, eval = has_not}
not_wrapper(x, contrast = "pcwsConstMean", seed = 1)
```

```{r mosum, eval = has_mosum}
mosum_wrapper(x)
mosum_wrapper(x3, multiscale = TRUE)
```

```{r idetect, eval = has_idetect}
idetect_wrapper(x, seed = 1)
```

```{r tguh, eval = has_breakfast}
tguh_wrapper(x)
```

## The 0.4.0 wave

**Multiscale inference with confidence.** SMUCE [@frick2014smuce] and its
heteroskedastic extension HSMUCE [@pein2017hsmuce] control the probability
of over-estimating the number of changes and return a confidence interval
for every changepoint location (`ci_lower`/`ci_upper`):

```{r smuce, eval = has_stepR}
res_smuce <- smuce_wrapper(x, alpha = 0.5)
tidy(res_smuce)
```

**Change in slope.** CPOP performs exact penalised estimation of a
continuous piecewise-linear mean [@fearnhead2019cpop; @fearnhead2024cpop]:

```{r cpop, eval = has_cpop}
y_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200)
res_cpop <- cpop_wrapper(y_slope)
tidy(res_cpop)
```

```{r cpop-fallback, include = FALSE, eval = !has_cpop}
# y_slope is reused by the segmented example below
y_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200)
```

**Bayesian detection.** `bcp_wrapper()` implements the Barry–Hartigan
product-partition model [@barry1993bayesian; @erdman2007bcp] and reports a
posterior probability per location; `bocpd_wrapper()` runs Bayesian online
changepoint detection over the run-length posterior [@adams2007bocpd];
`beast_wrapper()` wraps the BEAST Bayesian model-averaging ensemble
[@zhao2019beast]:

```{r bcp, eval = has_bcp}
res_bcp <- bcp_wrapper(x, seed = 1)
tidy(res_bcp)
```

```{r bocpd, eval = has_ocp}
res_bocpd <- bocpd_wrapper(x)
tidy(res_bocpd)
```

```{r beast, eval = has_rbeast}
res_beast <- beast_wrapper(x, seed = 1)
tidy(res_beast)
```

**Sequential and kernel nonparametrics.** `cpm_wrapper()` runs
distribution-free sequential tests and reports when each change would have
been *detected* in a stream [@ross2015cpm]; `kcp_wrapper()` applies kernel
change point analysis to running statistics (mean, variance,
autocorrelation, correlation) [@arlot2019kernel; @cabrieto2018kcprs];
`npmojo_wrapper()` detects distributional changes under serial dependence
[@mcgonigle2023npmojo]:

```{r cpm, eval = has_cpm}
tidy(cpm_wrapper(x, cpm_type = "Mann-Whitney"))
```

```{r kcp, eval = has_kcprs}
tidy(kcp_wrapper(x, running_stat = "mean", nperm = 100, seed = 1))
```

```{r npmojo, eval = has_cptnonpar}
tidy(npmojo_wrapper(x))
```

**Robustness to drift and dependence.** DeCAFS detects abrupt changes when
the signal also drifts and the noise is autocorrelated [@romano2022decafs];
self-normalised segmentation avoids long-run variance estimation entirely
[@zhao2022snseg]; EnvCpt only reports changepoints when a
changepoint model beats trend and autoregressive alternatives
[@beaulieu2018envcpt]:

```{r decafs, eval = has_decafs}
tidy(decafs_wrapper(x))
```

```{r sn, eval = has_snseg}
tidy(sn_wrapper(x3, parameter = "mean"))
```

```{r envcpt, eval = has_envcpt}
res_env <- envcpt_wrapper(x, models = c("mean", "meancpt", "trendcpt"))
tidy(res_env)
res_env$penalty$type   # which model won
```

**High-dimensional and multivariate.** `inspect_wrapper()` finds sparse
mean changes by projection [@wang2018inspect]; `ocd_wrapper()` monitors a
high-dimensional stream online [@chen2022ocd]; `geomcp_wrapper()` maps each
observation to a distance and an angle and segments both
[@grundy2020geomcp]; multivariate `ecp` input flows through `cpt_detect()`
unchanged.

```{r hd-data}
set.seed(1)
X <- cbind(a = c(rnorm(80), rnorm(80, 4)),
           b = c(rnorm(80), rnorm(80, -3)),
           c = rnorm(160))
```

```{r inspect, eval = has_inspect}
res_hd <- inspect_wrapper(X)
tidy(res_hd)
```

```{r ocd, eval = FALSE}
# Online high-dimensional detection; Monte Carlo threshold calibration makes
# this the slowest wrapper, so it is shown but not run here.
res_ocd <- ocd_wrapper(X, mc_reps = 100)
tidy(res_ocd)
```

```{r geomcp, eval = has_geomcp}
tidy(geomcp_wrapper(X))
```

**Regression structure.** `strucchange_wrapper()` dates Bai–Perron breaks
with confidence intervals [@bai1998estimating; @bai2003computation;
@zeileis2002strucchange] and accepts either a bare series or a formula;
`segmented_wrapper()` fits continuous broken-line regressions with
breakpoint standard errors [@muggeo2003segmented; @muggeo2008segmented]:

```{r strucchange, eval = has_struc}
tidy(strucchange_wrapper(x))
```

```{r segmented, eval = has_segmented}
tidy(segmented_wrapper(y_slope, npsi = 1, seed = 1))
```

**The modern PELT family.** `fastcpd_wrapper()` exposes the *fastcpd*
engine — mean, variance, mean-and-variance, and AR/ARMA/GARCH model
changepoints under one interface [@li2024fastcpd]:

```{r fastcpd, eval = has_fastcpd}
tidy(fastcpd_wrapper(x, family = "mean"))
```

# The penalty path

Rather than guessing one penalty, `cpt_crops()` computes *every* optimal
segmentation over a penalty interval (the CROPS algorithm, via
*changepoint* [@killick2014changepoint]) and returns a `ggcpt_path` with
its own `print()`, `tidy()`, and three plots:

```{r crops}
path <- cpt_crops(x3)
path
tidy(path)
```

```{r crops-elbow}
autoplot(path)                          # the cost elbow
```

```{r crops-path}
autoplot(path, type = "path")           # changepoints vs penalty
```

```{r crops-seg}
autoplot(path, type = "segmentations")  # the candidate models themselves
```

# The visualization layer

`autoplot()` renders any `ggcpt`. Options: `show_segments` (fitted segment
means), `show_fit` (the engine's own fitted signal, where provided —
SMUCE, DeCAFS, CPOP, segmented, bcp, BEAST), `show_ci`
(changepoint-location confidence intervals, where provided — SMUCE/HSMUCE,
strucchange, segmented), `show_points`/`show_line`, and the `cptline_*`
styling arguments (`cptline_color`, `cptline_alpha`, `cptline_type`,
`cptline_linewidth`).

```{r autoplot}
autoplot(res, show_segments = TRUE, cptline_color = "firebrick",
         cptline_type = "dashed", cptline_linewidth = 0.8)
```

```{r autoplot-ci, eval = has_stepR}
autoplot(res_smuce, show_ci = TRUE, show_fit = TRUE)
```

Multivariate results facet automatically:

```{r autoplot-mv, eval = has_inspect, fig.height = 6}
autoplot(res_hd)
```

The composable layers work inside any ggplot pipeline:
`geom_changepoint()` draws vertical rules, `geom_cpt_segment()` draws
segment levels, `geom_cpt_ci()` draws horizontal interval whiskers, and
`stat_changepoint()` runs detection *inside* the plot. `theme_ggcpt()` is a
publication theme and `annotate_segments()` shades alternating segments.

```{r geoms}
cp_tbl <- tidy(res)
df <- data.frame(index = seq_along(x), value = x)

ggplot(df, aes(index, value)) +
  annotate_segments(cp = cp_tbl$cp, n = length(x)) +
  geom_line() +
  geom_changepoint(data = cp_tbl, aes(xintercept = cp), color = "red") +
  geom_cpt_segment(
    data = res$segments,
    aes(x = start, xend = end, y = param_estimate, yend = param_estimate),
    inherit.aes = FALSE, color = "darkred", linewidth = 1
  ) +
  theme_ggcpt()
```

```{r stat}
ggplot(df, aes(index, value)) +
  geom_line() +
  stat_changepoint(method = "pelt", color = "blue")
```

```{r geom-ci, eval = has_stepR}
ci_tbl <- tidy(res_smuce)
ci_tbl$y_pos <- min(x) - 1
ggplot(df, aes(index, value)) +
  geom_line(color = "grey60") +
  geom_cpt_ci(
    data = ci_tbl,
    aes(y = y_pos, xmin = ci_lower, xmax = ci_upper),
    width = 0.6, color = "blue", inherit.aes = FALSE
  ) +
  geom_changepoint(data = ci_tbl, aes(xintercept = cp), color = "blue")
```

`ggcptplot()` and `ggecpplot()` are the original one-call plots for the two
classical engines:

```{r original-plots}
ggcptplot(x, change_in = "mean", cp_method = "PELT")
```

```{r ggecpplot}
ggecpplot(x, algorithm = "divisive", seed = 1)
```

The Bayesian engines get the field's signature displays:
`ggcpt_posterior()` shows the posterior mean over the series and the
per-location changepoint probability; `ggcpt_runlength()` shows the BOCPD
run-length posterior as a heatmap.

```{r posterior, eval = has_bcp, fig.height = 6}
ggcpt_posterior(res_bcp)
```

```{r runlength, eval = has_ocp}
ggcpt_runlength(res_bocpd)
```

Finally, `ggcpt_interactive()` turns any result (or ggplot built from one)
into a `plotly` widget with values on hover:

```{r interactive, eval = FALSE}
ggcpt_interactive(res)   # requires plotly; opens an htmlwidget
```

# Comparing methods

`ggcpt_compare()` runs several detectors on the same series and renders
them faceted (default) or overlaid; `ggcpt_compare_table()` returns the
tidy union. Both honour `future::plan()` for parallel execution.

```{r compare, fig.height = 6}
ggcpt_compare(x, methods = c("pelt", "binseg", "amoc"))
```

```{r compare-overlay}
ggcpt_compare(x, methods = c("pelt", "binseg"), layout = "overlay")
```

```{r compare-table}
ggcpt_compare_table(x, methods = c("pelt", "binseg", "amoc"))
```

# Batch detection and stability

`cpt_batch()` runs one detector over many series (matrix, data frame, or
list of vectors) and returns a tidy `ggcpt_batch` tibble with list-columns:

```{r batch}
XB <- cbind(shifted = x, pure_noise = rnorm(200))
batch <- cpt_batch(XB, method = "pelt")
batch
tidy(batch)
autoplot(batch)
```

`cpt_stability()` quantifies how fragile a segmentation is: it resamples
residuals within the fitted segments, re-runs the detector, and reports the
re-detection frequency at every location — a model-agnostic confidence
signal for engines with no native intervals:

```{r stability}
st <- cpt_stability(x, method = "pelt", B = 30, seed = 1)
st
autoplot(st)
```

# Evaluation against ground truth

`cpt_metrics()` scores predictions against a known truth: precision, recall
and F1 under one-to-one matching, the covering metric and adjusted Rand
index in the conventions of @van2020evaluation, Hausdorff distance,
annotation error, and matched MAE/RMSE. `cpt_metrics_annotated()` averages
over multiple annotators, and `ggcpt_eval()` draws the agreement (true
positives, false positives, and misses):

```{r metrics}
truth <- c(100)
pred <- tidy(res)$cp
cpt_metrics(pred, truth, n = length(x), margin = 5)
cpt_metrics_annotated(pred, list(c(100), c(101), c(99)), n = length(x))
```

```{r eval-plot}
ggcpt_eval(pred, truth, x, margin = 5)
```

# Simulation and canonical signals

`cpt_simulate()` (alias `rcpt()`) generates series with known changepoints
in mean, variance, both, or slope, under Gaussian, Student-t, AR(1), or
random-walk noise; the truth is carried in attributes. Five canonical test
signals from the literature ship ready-made: `signal_blocks()` (the
Donoho–Johnstone blocks signal [@donoho1994ideal]), `signal_fms()`,
`signal_mix()`, `signal_teeth()`, and `signal_stairs()`.

```{r simulate}
sim <- cpt_simulate(300, changepoints = c(100, 200), change_in = "mean",
                    params = c(0, 5, 1), seed = 1)
attr(sim, "true_changepoints")
sim2 <- rcpt(300, changepoints = 150, params = c(0, 3), seed = 2)

blocks <- signal_blocks(1024, seed = 1)
fms    <- signal_fms(500, seed = 1)
mix    <- signal_mix(500, seed = 1)
teeth  <- signal_teeth(400, seed = 1)
stairs <- signal_stairs(500, seed = 1)
```

```{r blocks-plot}
ggplot(blocks, aes(index, value)) +
  geom_line(color = "grey50") +
  geom_vline(xintercept = attr(blocks, "true_changepoints"),
             color = "blue", linewidth = 0.3) +
  labs(title = "The blocks test signal with its true changepoints")
```

# Citing the methodology

`cpt_cite()` returns the verified reference(s) behind a result or a method
name, so a write-up can cite the right paper without leaving R:

```{r cite}
cpt_cite("pelt")
cpt_cite(res)
```

# Closing note

This tour visited every exported function once. For the framework's design,
the mathematics of the wrapped methods, and worked analyses, see
`vignette("introduction", package = "ggchangepoint")`; for method
comparison and accuracy evaluation in depth, see
`vignette("comparison", package = "ggchangepoint")`.

# References
