---
title: "Inference, Selection and Diagnostics"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference, Selection and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint inference and diagnostic plot"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_ggcpt())

has_nsp   <- requireNamespace("nsp", quietly = TRUE)
has_mosum <- requireNamespace("mosum", quietly = TRUE)
has_wbs   <- requireNamespace("wbs", quietly = TRUE)
has_cvcp  <- requireNamespace("crossvalidationCP", quietly = TRUE)
has_infl  <- requireNamespace("changepoint.influence", quietly = TRUE)
```

Detection gives you locations. This vignette is about the four questions
that come next, and that a bare list of locations cannot answer:

1. **How sure are we about *where*?** — `cpt_confint()`, `nsp_wrapper()`.
2. **Is the change real at all?** — `cpt_test()`, and why the answer is
   harder than it looks.
3. **How many changes are there?** — `cpt_select()`.
4. **What is driving this answer?** — `cpt_influence()`,
   `cpt_sensitivity()`, `cpt_statistic()`.

```{r data}
set.seed(2026)
x <- c(rnorm(120), rnorm(120, 3), rnorm(120, 0.5))
fit <- cpt_detect(x, method = "pelt")
fit
```

# 1. Where could the changepoint be?

## Four provenances, one contract

Only a handful of engines ship an interval of their own. `cpt_confint()`
covers the rest, and — because the four routes mean genuinely different
things — reports which one it used in a `source` column rather than
presenting them as interchangeable.

```{r confint-bootstrap}
cpt_confint(fit, method = "bootstrap", B = 100, seed = 1)
```

The bootstrap route resamples residuals *within* the fitted segments,
re-runs the detector, and takes quantiles of the re-detected location. It
measures the sampling variability of the procedure conditional on the
fitted segmentation. It is model-agnostic and available for every engine;
it is not exact.

When the engine has its own interval, `method = "auto"` uses it:

```{r confint-native, eval = requireNamespace("stepR", quietly = TRUE)}
sm <- smuce_wrapper(x)
cpt_confint(sm)
```

Those are SMUCE's *simultaneous* confidence sets — a different object from
a bootstrap frequency band, which is exactly why the `source` column
exists.

## A changepoint that is an interval

Narrowest Significance Pursuit [@fryzlewicz2024nsp] inverts the usual
framing. Rather than estimating locations and then asking whether they are
real, it returns a set of intervals, each of which contains at least one
changepoint, with the guarantee holding **globally** across all the
intervals at level $\alpha$. The coverage is exact and finite-sample.

```{r nsp, eval = has_nsp}
res_nsp <- nsp_wrapper(x, alpha = 0.1, M = 200, seed = 1)
cpt_regions(res_nsp)
```

```{r nsp-plot, eval = has_nsp, fig.alt = "Series with shaded vertical bands marking NSP significance regions"}
autoplot(res_nsp)
```

Read the widths. A narrow band is a sharply located change; a wide one is
the honest statement that the data pin the change down only loosely. Note
what the result does **not** claim: NSP produces no point estimate, so the
`cp` column holds the interval midpoint, is labelled `region_midpoint` in
`cp_source`, and says so when the object prints. The region is the
inferential object.

The self-normalised and autoregressive variants keep the guarantee under
heavy tails, heteroscedasticity and serial dependence:

```{r nsp-ar, eval = has_nsp}
y_ar <- as.numeric(stats::arima.sim(list(ar = 0.6), 300)) +
  rep(c(0, 3), each = 150)
cpt_regions(nsp_wrapper(y_ar, variant = "ar", ord = 1, M = 200, seed = 1))
```

# 2. Is the change real?

This is where changepoint analysis is most often done wrong. Testing a
change at a location that was *chosen because the data looked like it
changed there* is circular, and the resulting p-values are
anti-conservative, often severely.

`cpt_test()` never hides that. It uses the engine's own test where one
exists and an explicitly unadjusted two-sample test where none does, and
the `selection_adjusted` column records which you got:

```{r test}
suppressWarnings(cpt_test(fit))
```

```{r test-native, eval = requireNamespace("strucchange", quietly = TRUE)}
cpt_test(strucchange_wrapper(x))
```

If you need a guarantee that survives selection, the route is NSP (above)
or `cpt_confint(method = "nsp")`, which maps each detected changepoint to
the narrowest region covering it and returns `NA` for one that no region
covers — because "no region supports this at level $\alpha$" is a finding,
not a missing value.

```{r confint-nsp, eval = has_nsp}
cpt_confint(fit, method = "nsp", level = 0.9, seed = 1)
```

# 3. How many changepoints?

`cpt_crops()` draws the penalty path. `cpt_select()` chooses from it, over
one candidate ladder shared by every criterion — so "BIC says 2,
cross-validation says 3" is a comparison of criteria and not of two
different searches.

```{r select}
sel <- cpt_select(x, criterion = "mbic", k_max = 8)
sel
```

`criterion = "mbic"` is the segment-length modified BIC of
@zhang2007mbic — the real one, which reads the segment lengths and so
cannot be expressed by `cpt_penalty()`'s function of $n$ and $k$ alone.

```{r select-criterion, fig.alt = "Criterion value against the number of changepoints, with the chosen model highlighted"}
autoplot(sel)
```

The ladder plot is more informative than the criterion curve, because it
shows what each candidate actually *is*:

```{r select-ladder, fig.height = 7, fig.alt = "Small multiples showing how the segmentation coarsens as the number of changepoints falls"}
autoplot(sel, plot_type = "ladder", max_facets = 6)
```

Cross-validation is the criterion with a consistency proof
[@zou2020copps]:

```{r select-cv, eval = has_cvcp}
cpt_select(x, criterion = "cv", k_max = 8)$k
```

A note on AIC: its $2k$ penalty does not grow with $n$, so it over-selects
and will usually take every rung the ladder offers. It is available
because people ask for it and because seeing the curve is instructive, not
because it is a good default.

# 4. What is driving the answer?

## Which observation

`cpt_stability()` answers "would I find this again?". `cpt_influence()`
asks the sharper question — which single observation, if perturbed,
changes the segmentation — following @wilms2022influence.

```{r influence}
inf <- cpt_influence(fit, engine = if (has_infl) "auto" else "recompute",
                     subset = if (has_infl) NULL else seq(1, 360, by = 6))
inf
```

```{r influence-overview, fig.alt = "Series with each observation sized and coloured by its influence on the segmentation"}
autoplot(inf)
```

```{r influence-map, eval = has_infl, fig.alt = "Influence map: perturbed observation on the horizontal axis, position on the vertical, segment-parameter shift as fill"}
autoplot(inf, plot_type = "map")
```

`cpt_leverage()` ranks the observations:

```{r leverage}
head(cpt_leverage(inf), 5)
```

## Which setting

```{r sensitivity, fig.height = 7, fig.alt = "One facet per penalty setting, each showing the changepoints that setting finds"}
sens <- cpt_sensitivity(x, method = "pelt",
                        over = list(penalty = c(2, 8, 20, 60)))
sens
autoplot(sens)
```

This is the direct answer to the commonest reviewer question about a
changepoint analysis, and it is worth running before the analysis is
written up rather than after it is questioned.

## What the detector computed

Every detector evaluates something at every location and then keeps only
the argmax. Three accessors give the discarded object back.

```{r statistic, eval = has_mosum, fig.alt = "Two-panel display: the series above, the MOSUM statistic against its threshold below"}
res_mosum <- cpt_detect(x, method = "mosum")
autoplot(res_mosum, type = "statistic")
```

The scale-space view answers a question a single-bandwidth fit cannot:
*at which resolutions does this feature exist?* A change visible only at a
wide bandwidth is a slow shift; one visible only at a narrow bandwidth is a
spike.

```{r scale-space, eval = has_mosum, fig.alt = "Heatmap of the MOSUM statistic by location and bandwidth, with accepted changepoints marked"}
ggcpt_scale_space(res_mosum, bandwidths = c(15, 30, 60, 90))
```

And the solution path shows the order in which candidates entered the
model, and how decisively each beat the next:

```{r path, eval = has_wbs, fig.alt = "Solution path: each candidate changepoint against the step at which it entered, with the proposing interval drawn"}
ggcpt_solution_path(cpt_detect(x, method = "wbs"), max_steps = 20)
```

An engine that exposes none of these says so, and names the ones that do:

```{r statistic-error, error = TRUE}
cpt_statistic(fit)
```

# Putting it together

```{r report}
cat(head(cpt_report(fit, session = FALSE), 20), sep = "\n")
```

# References
