---
title: "ggchangepoint: A Unified Tidy Interface for Changepoint Analysis in R"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggchangepoint: A Unified Tidy Interface for Changepoint Analysis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.alt = "ggchangepoint plot of a time series with detected changepoints"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_light())

# Optional (Suggests) engines: gate the chunks that need them so the
# vignette builds on any installation.
has_stepR       <- requireNamespace("stepR", quietly = TRUE)
has_cpop        <- requireNamespace("cpop", quietly = TRUE)
has_bcp         <- requireNamespace("bcp", quietly = TRUE)
has_ocp         <- requireNamespace("ocp", quietly = TRUE)
has_fpop        <- requireNamespace("fpop", quietly = TRUE)
has_wbs         <- requireNamespace("wbs", quietly = TRUE)
has_not         <- requireNamespace("not", quietly = TRUE)
has_mosum       <- requireNamespace("mosum", quietly = TRUE)
has_idetect     <- requireNamespace("IDetect", quietly = TRUE)
has_breakfast   <- requireNamespace("breakfast", quietly = TRUE)
has_cpm         <- requireNamespace("cpm", quietly = TRUE)
has_decafs      <- requireNamespace("DeCAFS", quietly = TRUE)
has_inspect     <- requireNamespace("InspectChangepoint", quietly = TRUE)
has_strucchange <- requireNamespace("strucchange", quietly = TRUE)
has_segmented   <- requireNamespace("segmented", quietly = TRUE)
has_fastcpd     <- requireNamespace("fastcpd", quietly = TRUE)
has_envcpt      <- requireNamespace("EnvCpt", quietly = TRUE)
```

# Abstract

**ggchangepoint** is an R package that provides a unified, tidy interface to
changepoint detection across the methodological spectrum. It introduces a
single S3 result class, `ggcpt`, with `broom`-style methods (`tidy()`,
`glance()`, `augment()`) [@robinson2017broom], a central dispatcher
`cpt_detect()` covering 31 detection methods across five algorithmic
families, and native `ggplot2` [@wickham2016ggplot2] visualisation through
`autoplot()` and a set of composable geoms. Where a method quantifies its
uncertainty — the simultaneous confidence intervals of SMUCE
[@frick2014smuce], the break-date intervals of Bai–Perron
[@bai1998estimating], the posterior probabilities of Bayesian detectors
[@barry1993bayesian; @adams2007bocpd] — the result object carries that
uncertainty and the plotting layer can draw it. The package further supplies
a penalty-path diagnostic (CROPS), batch detection over panels of series,
bootstrap stability diagnostics, accuracy metrics aligned with modern
benchmarking conventions [@van2020evaluation], ground-truth simulation, and
per-method citations. This article describes the statistical background, the
design of the package, and each method family in turn, with worked examples
throughout.

# Introduction

Changepoint analysis — locating the instants at which the stochastic
behaviour of an ordered sequence changes — is one of the oldest problems in
statistics, dating back at least to the quality-control charts of
@page1954continuous, and one of its most active: modern surveys catalogue
dozens of families of methods [@truong2020selective;
@aminikhanghahi2017survey]. Its applications span virtually every domain
that produces sequential data, including genomics [@picard2005statistical],
finance [@athey2022detecting], climate science [@haslett1989space], and
signal processing [@lavielle2005using].

The R ecosystem mirrors this breadth. Penalised optimal partitioning lives
in **changepoint** [@killick2014changepoint] and **fpop**
[@maidstone2017optimal]; wild binary segmentation in **wbs** and
**breakfast** [@fryzlewicz2014wild; @fryzlewicz2020detecting]; multiscale
inference in **stepR** [@frick2014smuce]; Bayesian analysis in **bcp**
[@erdman2007bcp] and **ocp** [@adams2007bocpd]; structural breaks in
**strucchange** [@zeileis2002strucchange]; and so on. Each of these packages
is excellent at what it does, and each returns a different object, follows a
different indexing convention, and draws (or does not draw) its own plots.

An analyst who wants to *compare* a PELT segmentation with a Bayesian
posterior and a multiscale confidence set — a routine task in applied work —
must therefore learn several APIs, reconcile several conventions, and write
custom plotting code for each. **ggchangepoint** removes that friction. Its
design goals are:

1. **One vocabulary.** A single front door, `cpt_detect(x, method,
   change_in, penalty, ...)`, whose arguments mean the same thing for every
   engine.
2. **One result type.** Every detector returns a `ggcpt` object with a
   stable tidy contract, whatever the upstream engine returned.
3. **One rendering path.** Every result — point estimates, confidence
   intervals, fitted signals, posteriors, penalty paths — draws with
   `autoplot()` and extends with ordinary `ggplot2` layers.
4. **Wrap, don't reinvent.** All detection is delegated to the
   peer-reviewed upstream engines; optional engines live in `Suggests` and
   are loaded only when requested.

# The changepoint problem

Let $y_{1:n} = (y_1, \dots, y_n)$ be an ordered sequence. A segmentation
with $m$ changepoints is an ordered set $\tau_{1:m}$ with
$0 = \tau_0 < \tau_1 < \dots < \tau_m < \tau_{m+1} = n$, partitioning the
data into $m + 1$ segments $y_{(\tau_{i-1}+1):\tau_i}$. Throughout the
package a changepoint $\tau$ is reported as the **last index of the left
segment** (the convention of the **changepoint** package); results from
engines using the opposite convention are normalised on the way in, and the
convention is recorded on every result object.

## Penalised cost minimisation

The classical formulation chooses the segmentation minimising a penalised
cost,
$$
\min_{m,\ \tau_{1:m}} \; \sum_{i=1}^{m+1}
  \mathcal{C}\!\left(y_{(\tau_{i-1}+1):\tau_i}\right) \;+\; \beta m,
$$
where $\mathcal{C}$ is a segment cost (for a change in mean under Gaussian
noise, the residual sum of squares; more generally twice the negative
maximised log-likelihood) and $\beta > 0$ penalises each additional
changepoint. Common choices of $\beta$ are AIC, BIC/SIC
[@yao1988estimating], and modified BIC variants. Solving this optimisation
exactly by dynamic programming costs $O(n^2)$; **PELT**
[@killick2012pelt] prunes candidate changepoints to achieve linear expected
cost while remaining exact, and **FPOP** [@maidstone2017optimal] achieves
comparable speed by functional pruning.

## Search-based and multiscale methods

A complementary family locates changepoints by scanning test statistics.
**Binary segmentation** [@scott1974cluster; @vostrikova1981detecting]
recursively splits the series at the maximal CUSUM statistic; **wild binary
segmentation** [@fryzlewicz2014wild] and its successor WBS2
[@fryzlewicz2020detecting] draw random subintervals so that short segments
are not masked; **narrowest-over-threshold** (NOT)
[@baranowski2019narrowest] favours the narrowest interval on which the
contrast exceeds a threshold, which generalises cleanly to changes in slope;
**MOSUM** [@eichinger2018mosum] scans a moving-sum statistic at a fixed
bandwidth; Isolate–Detect [@anastasiou2022idetect] isolates each
changepoint in an expanding interval; and TGUH [@fryzlewicz2022tail]
performs a tail-greedy bottom-up merge. **SMUCE** [@frick2014smuce]
occupies a special place: it estimates the step function with the fewest
jumps that passes a *simultaneous multiscale test* at level $\alpha$, and in
doing so delivers confidence intervals for every changepoint location —
uncertainty statements most competitors cannot make. HSMUCE
[@pein2017hsmuce] extends this to heterogeneous noise.

## Beyond the mean

Changes need not be in the mean: the package's `change_in` argument spans
variance, mean-and-variance, slope, and full distributional change.
Nonparametric engines (energy statistics [@matteson2014nonparametric],
nonparametric cost functions [@haynes2017computationally], kernel running
statistics [@arlot2019kernel; @cabrieto2018kcprs], joint characteristic
functions [@mcgonigle2023npmojo], self-normalisation [@zhao2022snseg])
detect distributional change without likelihood assumptions; Bayesian
engines [@barry1993bayesian; @adams2007bocpd; @zhao2019beast] return
posteriors instead of point sets; high-dimensional engines
[@wang2018inspect; @chen2022ocd; @grundy2020geomcp] aggregate evidence
across coordinates; and regression engines [@bai1998estimating;
@muggeo2003segmented] date breaks in model coefficients. The tour below
visits each family with runnable code.

# Design of the package

## The `ggcpt` result contract

Every detector returns an object of class `ggcpt` containing:

- `changepoints`: a tibble with one row per changepoint. Columns `cp`
  (location, "left" convention) and `cp_value` (the data value at `cp`) are
  always present; engines add `ci_lower`/`ci_upper` (SMUCE, HSMUCE,
  strucchange, segmented), `posterior_prob` (bcp, BEAST),
  `detection_time` (CPM), `strength` (inspect), `declared_at` (ocd), or
  `mapping` (geomcp) when they have more to say.
- `segments`: a tibble of the induced segments (`seg_id`, `start`, `end`,
  `n`, `param_estimate`).
- `data`: the analysed series as a tibble (`index`, `value`), plus a
  `fitted` column when the engine estimates a signal (SMUCE, DeCAFS, CPOP,
  segmented, bcp, BEAST).
- `method`, `change_in`, `penalty` (a `list(type, value)` descriptor),
  `cp_convention` (always `"left"`), `runtime` (seconds, measured by
  `cpt_detect()`), and `fit` (the untouched upstream object, for experts).

Multivariate results additionally carry a `data_wide` tibble with one column
per coordinate, which `autoplot()` renders as faceted small-multiples.

## Tidy methods and the plotting layer

The class implements the full complement of generics R users expect:

```{r contract}
set.seed(2022)
x <- c(rnorm(100, 0, 1), rnorm(100, 10, 1))
res <- cpt_detect(x, method = "pelt", change_in = "mean")
res
tidy(res)
glance(res)
head(augment(res))
```

`autoplot()` draws the series, the changepoint rules, and — on request —
the fitted segment means (`show_segments`), the engine's fitted signal
(`show_fit`), and changepoint-location confidence intervals (`show_ci`):

```{r contract-plot}
autoplot(res, show_segments = TRUE)
```

Composable layers (`geom_changepoint()`, `geom_cpt_segment()`,
`geom_cpt_ci()`, `stat_changepoint()`), a theme (`theme_ggcpt()`), and
segment shading (`annotate_segments()`) let the same results be built into
bespoke graphics; `summary()`, `as_tibble()`, `as.data.frame()`,
`format()`, and `plot()` complete the S3 surface.

## Design principles

The package follows seven principles, unchanged since 0.2.0:

- **P1 — Wrap, don't reinvent**: bind to peer-reviewed CRAN engines.
- **P2 — Tidy in, tidy out**: stable column names across all methods.
- **P3 — ggplot2 all the way down**: every result renders and extends.
- **P4 — One vocabulary**: `x`, `method`, `change_in`, `penalty`, `...`.
- **P5 — Progressive disclosure**: beginners call `cpt_detect()` +
  `autoplot()`; experts reach the upstream fit via `$fit`.
- **P6 — No surprises**: the 0.1.0 functions still work unchanged.
- **P7 — Document everything you ship**: every export is introduced in the
  README and a vignette.

Release 0.4.0 adds an eighth: **P8 — carry the uncertainty**. Where the
method quantifies uncertainty, the `ggcpt` object records it and
`autoplot()` can draw it.

# The unified dispatcher

`cpt_detect()` dispatches by method name; `cpt_methods()` reports every
method the package knows, its engine, what it can detect, and whether the
engine is installed:

```{r methods-table}
cpt_methods()
```

Requests are validated against this capability matrix: asking a mean-only
engine for a variance change is an error with the legal alternatives named —
never a silent substitution. Univariate methods likewise refuse multi-column
input rather than flattening it.

Penalty semantics differ across engines, and `cpt_penalty()` documents and
constructs the standard values:

```{r penalty}
cpt_penalty("BIC", n = 200)
cpt_penalty("MBIC", n = 200)
cpt_penalty("Hannan-Quinn", n = 200)
```

Character penalties (`"MBIC"`, `"BIC"`, ...) pass through to the
`changepoint`-family engines natively and are resolved to numeric values for
functional-pruning engines (`fpop`, `cpop`, `decafs`); search-based engines
(WBS, NOT, MOSUM, ...) use their own model-selection criteria and ignore the
argument.

# A tour of the method families

Throughout, we use simulated series with known truth so results can be
checked by eye. The simulator and canonical test signals
(`signal_blocks()` — the Donoho–Johnstone blocks signal [@donoho1994ideal] —
`signal_fms()`, `signal_teeth()`, `signal_stairs()`, `signal_mix()`) are
described in the comparison vignette.

```{r tour-data}
set.seed(2026)
x_mean  <- c(rnorm(100), rnorm(100, 4))                # mean shift at 100
x_multi <- c(rnorm(100), rnorm(100, 3), rnorm(100, -1)) # shifts at 100, 200
x_slope <- cumsum(c(rep(0.4, 100), rep(-0.3, 100))) + rnorm(200) # kink at 100
```

## Penalised and optimal partitioning

PELT [@killick2012pelt], binary segmentation [@scott1974cluster], segment
neighbourhoods, and AMOC come from the **changepoint** package; FPOP
[@maidstone2017optimal] from **fpop**:

```{r penalised}
tidy(cpt_detect(x_multi, method = "pelt"))
tidy(cpt_detect(x_multi, method = "binseg"))
```

```{r penalised-fpop, eval = has_fpop}
tidy(cpt_detect(x_multi, method = "fpop"))
```

The Achilles heel of penalised methods is the choice of $\beta$. Rather
than committing to one value, `cpt_crops()` computes *every* optimal
segmentation as $\beta$ ranges over an interval — the CROPS algorithm of
Haynes, Eckley and Fearnhead (2017), as implemented by **changepoint** —
and turns penalty selection into a diagnostic:

```{r crops}
path <- cpt_crops(x_multi)
path
autoplot(path)
```

The elbow at the true number of changepoints (here 2) is the standard
reading. `autoplot(path, type = "segmentations")` shows the candidate
models themselves, and `autoplot(path, type = "path")` the map from penalty
to model size:

```{r crops-segmentations}
autoplot(path, type = "segmentations")
```

The modern **fastcpd** engine [@li2024fastcpd] brings the same penalised
formulation to a wide family of models — mean, variance, mean-and-variance,
and AR/ARMA/GARCH model changes — with sequential-gradient-descent speed:

```{r fastcpd, eval = has_fastcpd}
tidy(fastcpd_wrapper(x_multi, family = "mean"))
```

## Multiscale and search methods

The randomised and multiscale searchers are one wrapper call each:

```{r search-wbs, eval = has_wbs}
tidy(wbs_wrapper(x_multi, seed = 1))
```

```{r search-not, eval = has_not}
tidy(not_wrapper(x_multi, seed = 1))
```

```{r search-mosum, eval = has_mosum}
tidy(mosum_wrapper(x_multi))
```

```{r search-others, eval = has_idetect && has_breakfast}
tidy(idetect_wrapper(x_multi, seed = 1))
tidy(wbs2_wrapper(x_multi))
tidy(tguh_wrapper(x_multi))
```

SMUCE [@frick2014smuce] is the family's inferential flagship: its level
$\alpha$ bounds the probability of overestimating the number of
changepoints, and every location comes with a confidence interval, stored in
`ci_lower`/`ci_upper` and drawn by `show_ci = TRUE` as whiskers beneath the
series (the step fit is drawn by `show_fit = TRUE`):

```{r smuce, eval = has_stepR}
res_smuce <- smuce_wrapper(x_multi)
tidy(res_smuce)
autoplot(res_smuce, show_ci = TRUE, show_fit = TRUE)
```

For heterogeneous noise, `smuce_wrapper(x, family = "hsmuce")` (or
`cpt_detect(x, method = "hsmuce")`) runs HSMUCE [@pein2017hsmuce].

## Changes in slope

A kink in the trend is not a jump in the level, and running a mean-change
detector on a trending series over-detects notoriously. CPOP
[@fearnhead2019cpop; @fearnhead2024cpop] solves the change-in-slope problem
*exactly* under an $L_0$ penalty, returning a continuous piecewise-linear
fit:

```{r cpop, eval = has_cpop}
res_cpop <- cpop_wrapper(x_slope)
tidy(res_cpop)
autoplot(res_cpop, show_fit = TRUE)
```

NOT with its linear contrast [@baranowski2019narrowest] offers a
search-based alternative; the dispatcher routes
`cpt_detect(x, method = "not", change_in = "slope")` to it automatically:

```{r not-slope, eval = has_not}
tidy(cpt_detect(x_slope, method = "not", change_in = "slope"))
```

## Bayesian detection

The Barry–Hartigan product partition model [@barry1993bayesian], via the
**bcp** package [@erdman2007bcp], returns a *posterior probability of a
changepoint at every location* along with posterior segment means. Locations
clearing `prob_threshold` populate the changepoints tibble (with their
probabilities), and `ggcpt_posterior()` draws the classic two-panel
display:

```{r bcp, eval = has_bcp}
res_bcp <- bcp_wrapper(x_mean, seed = 2026)
tidy(res_bcp)
ggcpt_posterior(res_bcp)
```

Bayesian *online* changepoint detection [@adams2007bocpd] instead tracks
the posterior over the current **run length** — the time since the last
change — updating recursively as data arrive. Its signature graphic is the
run-length heatmap, where a change appears as the posterior mass collapsing
to zero:

```{r bocpd, eval = has_ocp}
res_bocpd <- bocpd_wrapper(x_mean)
tidy(res_bocpd)
ggcpt_runlength(res_bocpd)
```

A third Bayesian engine, BEAST [@zhao2019beast] via **Rbeast**, averages
over models rather than conditioning on one, and is wired as
`cpt_detect(x, method = "beast")` (or `beast_wrapper()`); it too reports
`posterior_prob` and renders with `ggcpt_posterior()`.

## Nonparametric and sequential detection

When no parametric form is trustworthy, the nonparametric cost approach of
**changepoint.np** [@haynes2017computationally] and the energy-statistics
E-Divisive of **ecp** [@matteson2014nonparametric; @james2014ecp] detect
general distributional change:

```{r nonparam}
set.seed(2022)
tidy(cpt_detect(x_mean, method = "np"))
tidy(cpt_detect(x_mean, method = "ecp", seed = 1))
```

The **cpm** package [@ross2015cpm] runs sequential two-sample tests
(Mann–Whitney, Mood, Lepage, Kolmogorov–Smirnov, ...) as an online monitor;
its results distinguish where a change *happened* (`cp`) from when it was
*detected* (`detection_time`), the inherent lag of sequential monitoring:

```{r cpm, eval = has_cpm}
tidy(cpm_wrapper(x_mean, cpm_type = "Mann-Whitney"))
```

Three further nonparametric engines are wired and worth knowing: kernel
change-point analysis on running statistics (`kcp_wrapper()`, engine
**kcpRS**), which detects changes in running means, variances,
autocorrelations, or correlations [@arlot2019kernel; @cabrieto2018kcprs];
NP-MOJO (`npmojo_wrapper()`, engine **CptNonPar**), which detects changes in
the marginal or lagged joint distribution while remaining valid under serial
dependence [@mcgonigle2023npmojo]; and self-normalised segmentation
(`sn_wrapper()`, engine **SNSeg**), which avoids long-run variance
estimation altogether and tests changes in means, variances,
autocorrelations, or bivariate correlations [@zhao2022snseg].

## Robustness to drift, autocorrelation, and model ambiguity

The most common failure of naive mean-change detection in practice is not a
subtle statistical issue: it is running a Gaussian-mean detector on data
whose baseline drifts or whose noise is autocorrelated, and reporting a
changepoint wherever the model is wrong. DeCAFS [@romano2022decafs] models
exactly this regime — abrupt changes superimposed on random-walk drift and
AR(1) noise — and separates the two:

```{r decafs, eval = has_decafs}
res_decafs <- decafs_wrapper(x_mean)
tidy(res_decafs)
autoplot(res_decafs, show_fit = TRUE)
```

EnvCpt [@beaulieu2018envcpt] attacks the same confusion by model
selection: it fits up to twelve competing descriptions — constant mean or
trend, with or without changepoints, with white or AR(1)/AR(2) noise — and
reports changepoints only if a changepoint model wins on information
criteria:

```{r envcpt, eval = has_envcpt}
res_env <- envcpt_wrapper(x_mean, models = c("mean", "meancpt", "trendcpt"))
glance(res_env)
```

The winning model's name is recorded in the penalty descriptor
(`penalty_type` above), so "no changepoints, it's just autocorrelation" is a
first-class answer.

## Multivariate and high-dimensional detection

Multivariate methods accept a matrix (rows are time points) directly. The
energy-statistics E-Divisive of **ecp** was built for this
[@matteson2014nonparametric]; for high-dimensional data whose change is
confined to a sparse subset of coordinates, `inspect` [@wang2018inspect]
finds an optimal sparse projection of the CUSUM matrix and reports the
projected evidence (`strength`). Multivariate results render as faceted
small-multiples with shared changepoint rules:

```{r inspect, eval = has_inspect}
set.seed(2026)
X <- cbind(a = c(rnorm(80), rnorm(80, 3)),
           b = c(rnorm(80), rnorm(80, -2)),
           c = rnorm(160))
res_hd <- inspect_wrapper(X)
tidy(res_hd)
autoplot(res_hd)
```

Two further engines complete the family: `geomcp_wrapper()` (engine
**changepoint.geo**) maps each observation to its distance from and angle to
a reference point and segments the two mapped series, catching magnitude and
orientation changes respectively [@grundy2020geomcp]; and `ocd_wrapper()`
(engine **ocd**) monitors a high-dimensional stream *online* with
worst-case detection-delay guarantees [@chen2022ocd], reporting both the
declaration time and (by construction of the wrapper) handling baseline
estimation and post-declaration restarts.

## Structural breaks in regression

Econometric practice dates breaks in regression coefficients. The
Bai–Perron estimator [@bai1998estimating; @bai2003computation], via
**strucchange** [@zeileis2002strucchange], returns break dates *with
confidence intervals*; called on a bare series it dates mean shifts, and
called with a formula it dates breaks in arbitrary regressions:

```{r strucchange, eval = has_strucchange}
res_bp <- strucchange_wrapper(x_mean)
tidy(res_bp)
autoplot(res_bp, show_ci = TRUE)
```

Where the regression function is continuous — a kink rather than a jump —
**segmented** [@muggeo2003segmented; @muggeo2008segmented] estimates
broken-line relationships with standard errors for the breakpoints:

```{r segmented, eval = has_segmented}
res_seg <- segmented_wrapper(x_slope, npsi = 1, seed = 1)
tidy(res_seg)
autoplot(res_seg, show_fit = TRUE, show_ci = TRUE)
```

# Beyond detection

## Batch detection over many series

Applied work rarely stops at one series. `cpt_batch()` runs a detector over
every column of a matrix or data frame (or every element of a list) and
returns a tidy tibble of results, with `future`-based parallelism available
via `future::plan()`:

```{r batch}
set.seed(2026)
panel <- cbind(shifted = x_mean, quiet = rnorm(200))
batch <- cpt_batch(panel, method = "pelt")
batch
tidy(batch)
autoplot(batch)
```

## Stability diagnostics

Most engines report a point set with no measure of its fragility.
`cpt_stability()` resamples residuals within the fitted segments, re-runs
the detector on each replicate, and reports how often each location is
re-detected — a cheap, model-agnostic confidence signal available for
*every* engine, including the many with no native intervals:

```{r stability}
st <- cpt_stability(x_mean, method = "pelt", B = 50, seed = 1)
st
autoplot(st)
```

## Evaluation, interactivity, and citations

When ground truth is known, `cpt_metrics()` computes precision, recall and
F1 under one-to-one matching, the covering metric, Hausdorff distance, and
adjusted Rand index, following the conventions of the modern benchmarking
literature [@van2020evaluation]; `ggcpt_eval()` draws the agreement, and
`ggcpt_compare()` juxtaposes methods. These are the subject of the
companion vignette `vignette("comparison", package = "ggchangepoint")`.

Any result renders as an interactive HTML widget with
`ggcpt_interactive(res)` (engine **plotly**, in `Suggests`); the static
`autoplot()` path is untouched.

Finally, because every method here is someone's published work, `cpt_cite()`
returns the reference(s) behind a result, so analyses can cite the right
paper without leaving R:

```{r cite}
cpt_cite("pelt")
```

# Discussion

ggchangepoint does not contribute a new detection algorithm; it contributes
a *surface*. The value of a common contract compounds with the number of
methods behind it: the same `tidy()` pipeline, the same plot, and the same
evaluation code now span penalised, multiscale, nonparametric, Bayesian,
high-dimensional, and regression-based detection — 31 methods in this
release. Methods whose engines are currently absent from CRAN
(graph-constrained gfpop [@hocking2020gfpop], robust segmentation
[@fearnhead2019changepoint], FOCuS, sparsified binary segmentation) are
tracked as *planned* in `cpt_methods()` and will slot into the same wrapper
pattern when their engines return.

Two practical notes. First, wrapped engines are run with sensible defaults,
but each wrapper forwards `...` to its engine, and the raw fit is always in
`$fit` — the package is a front door, not a cage. Second, detection quality
is the engines'; the package's own additions (metrics, stability, penalty
paths) are deliberately engine-agnostic, so conclusions drawn with them
transfer.

# Acknowledgements

This package stands on the shoulders of the authors of the wrapped engines
and of the R [@rcore], ggplot2 [@wickham2016ggplot2], and broom
[@robinson2017broom] projects.

# References
