---
title: "Streaming and Online Monitoring"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Streaming and Online Monitoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 72,
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggchangepoint monitoring plot"
)
library(ggchangepoint)
library(ggplot2)
theme_set(theme_ggcpt())

has_cpm <- requireNamespace("cpm", quietly = TRUE)
has_ocd <- requireNamespace("ocd", quietly = TRUE)
```

Every other vignette in this package segments a series that is already
complete. This one does not. A monitor consumes observations as they
arrive, and the question it answers is not *where* was the change but
**how long did you take to notice, and how often do you cry wolf**. Those
are different quantities, they are scored differently, and conflating them
is the commonest mistake made with online detectors — including by
retrospective plots that draw an alarm time as though it were a
changepoint location.

So a monitor is a different object from a segmentation. Five functions:

| Function | What it does |
|---|---|
| `cpt_monitor()` | build a stateful detector from a clean baseline |
| `cpt_update()` | push new observations through it |
| `alarms()` | the alarm log: when it fired, and on what statistic |
| `cpt_delay()` | score it — detection delay and false alarms |
| `cpt_replay()` | run a whole series through in one call |

# 1. A monitor, one batch at a time

Start from a baseline the detector may assume is in control. `edetector`
and `ocd` estimate their reference mean and scale from it; `cpm` derives
its own from a start-up period and so does not need one.

```{r build}
set.seed(2026)
baseline <- rnorm(200)
mon <- cpt_monitor("edetector", baseline = baseline, alpha = 0.002)
mon
```

Nothing has been monitored yet. Now feed it 120 observations from the same
distribution — the in-control case, where the right answer is silence:

```{r in-control}
set.seed(5201)
mon <- cpt_update(mon, rnorm(120))
mon
```

Then a change arrives. The mean shifts by two standard deviations at
observation 121 of the monitored stream, and the monitor sees it only one
observation at a time:

```{r change}
set.seed(5202)
mon <- cpt_update(mon, rnorm(80, mean = 2))
alarms(mon)
```

The idiom is `mon <- cpt_update(mon, new_obs)`: the monitor is returned
because it is stateful, and each call appends to the same alarm log. The
`time` column counts observations fed to the monitor, so the baseline is
not on the clock.

# 2. Scoring: delay, not location

`cpt_delay()` matches each alarm to the change it plausibly detected, and
counts the rest as false alarms.

```{r delay}
d <- cpt_delay(mon, truth = 121)
d
```

Read the three numbers that matter. **Delay** is how many observations
passed between the change and the first alarm after it. **False alarms**
are the alarms with no change behind them. **Average run length** is
observations per false alarm — the in-control cost of running the monitor,
and the quantity the thresholds are calibrated against.

Note what happened to the *second* alarm. Only the first alarm after a
change counts as the detection, so the later one is charged as a false
alarm even though the shift causing it is real. That is the right
accounting for a monitor whose job is to raise one flag per change, and
it is also a hint about `relearn`, which section 6 returns to.

`tidy()` and `glance()` give the same content as data:

```{r delay-tidy}
glance(d)
```

`cpt_metrics()` is the wrong tool here. It asks whether the *location* was
recovered, which a sequential procedure never claims: an alarm is
deliberately late, so scoring it as a location estimate marks a correct
detection wrong by exactly the delay.

```{r timeline, fig.alt = "Monitored series with dashed vertical lines at each alarm"}
autoplot(mon)
```

```{r delay-plot, fig.height = 3.5, fig.alt = "Bar chart of the detection delay at the true changepoint"}
autoplot(d)
```

# 3. Replaying a finished series

`cpt_replay()` does the update loop for you: it takes the leading
observations as the baseline, streams the rest, and returns the same
monitor object. This is how to study an online method's alarm timeline
retrospectively without pretending its alarms are changepoints.

```{r replay}
set.seed(5203)
stream <- c(rnorm(200), rnorm(200, mean = 2))
rep_e <- cpt_replay(stream, method = "edetector")
alarms(rep_e)
```

The baseline offset is applied automatically, so `truth` is on the
original clock of `stream`:

```{r replay-delay}
glance(cpt_delay(rep_e, truth = 200))
```

# 4. The three methods

## `edetector` — the default, governed by `alpha`

A mixture Shiryaev–Roberts e-detector for a sub-Gaussian shift
[@shin2023edetectors]. For each candidate shift the increment is a
likelihood ratio with unit mean under the null; the running statistics are
combined by **averaging**, which keeps $M_t - t$ a mean-zero martingale, so
optional stopping gives $E_\infty[\tau] \ge 1/\alpha$ — a finite-sample
lower bound on the in-control average run length, with no calibration run.
It is the one detector in this package implemented here rather than
wrapped, because no R package implements e-detectors; `print()` labels it
as such.

`deltas` is the set of shift sizes mixed over, in baseline standard
deviations, each taken in both directions. The mixture is a uniform
average, so adding a shift costs power at the ones already there rather
than inflating the false-alarm rate — and betting on the wrong one is
expensive. Eight replicates, median delay:

```{r deltas}
set.seed(2026)
delay_at <- function(deltas, shift, reps = 8, n = 200) {
  v <- vapply(seq_len(reps), function(i) {
    s <- c(rnorm(n), rnorm(n, mean = shift))
    cpt_delay(cpt_replay(s, method = "edetector", deltas = deltas),
              truth = n)$median_delay
  }, numeric(1))
  median(v, na.rm = TRUE)
}
data.frame(
  shift = c(0.75, 3),
  mixture_0.5_1_2 = c(delay_at(c(0.5, 1, 2), 0.75), delay_at(c(0.5, 1, 2), 3)),
  only_0.5 = c(delay_at(0.5, 0.75), delay_at(0.5, 3)),
  only_3 = c(delay_at(3, 0.75), delay_at(3, 3))
)
```

A detector that mixes only over a 3-sigma shift takes several times as
long to notice a 0.75-sigma one, while the mixture is close to the
best single choice at both sizes. That is the argument for mixing: it buys
robustness to not knowing the change size, at a small cost when you do.

## `cpm` — governed by `arl0`

`cpm`'s sequential change-point model [@ross2015cpm], with a
distribution-free statistic (`cpm_type`, `"Mann-Whitney"` by default) and
a threshold calibrated to a target in-control average run length.

```{r cpm, eval = has_cpm}
set.seed(5204)
mon_cpm <- cpt_monitor("cpm", arl0 = 500)
mon_cpm <- cpt_update(mon_cpm, rnorm(120))
mon_cpm <- cpt_update(mon_cpm, rnorm(80, mean = 2))
alarms(mon_cpm)
glance(cpt_delay(mon_cpm, truth = 121))
```

`cpm` does the threshold comparison inside the engine, so `threshold` is
`NA` in its alarm log rather than an invented number, and the `statistic`
column is not informative for it either. The alarm *time* is the whole
signal from `cpm`; the statistic trace is what `edetector` and `ocd`
provide and `cpm` does not.

## `alpha` and `arl0` are alternative parameterisations, not two knobs

`cpt_monitor()` exposes both because the two engines are calibrated in
different currencies, and each ignores the other's. **`edetector` honours
`alpha`** (its threshold is $1/\alpha$); **`cpm` honours `arl0`**. Setting
`arl0` on an e-detector, or `alpha` on `cpm`, changes nothing.

```{r alpha, warning = TRUE}
set.seed(5205)
ic <- rnorm(400)
vapply(c(0.05, 0.01, 0.001),
       function(a) nrow(alarms(cpt_replay(ic, method = "edetector",
                                          alpha = a))),
       numeric(1))
nrow(alarms(cpt_replay(ic, method = "edetector", arl0 = 5000)))
```

Three alphas, three false-alarm counts; then `arl0 = 5000` on the same
stream, which reproduces the default-`alpha` result exactly because the
argument is not read -- and `cpt_monitor()` says so, because an argument
that is accepted and then ignored is worth a warning rather than an
unchanged answer. The rough translation is $\mathrm{ARL}_0 \approx
1/\alpha$, so `alpha = 0.002` and `arl0 = 500` ask for comparable
strictness.

## `ocd` — multivariate only

`ocd`'s high-dimensional multiscale detector [@chen2022ocd] tracks a
projection of the whole vector, and **requires at least two
coordinates**. On a single series it does not fall back to a univariate
statistic; it stops, and says which methods do read one series:

```{r ocd-error, error = TRUE}
set.seed(5206)
cpt_monitor("ocd", baseline = rnorm(100))
```

Given a matrix with rows as time points it monitors all coordinates
jointly, and the alarms are shared across them:

```{r ocd, eval = has_ocd}
set.seed(11)
base_mv <- matrix(rnorm(200 * 3), ncol = 3)
mon_ocd <- cpt_monitor("ocd", baseline = base_mv, patience = 200,
                       mc_reps = 30)
stream_mv <- rbind(matrix(rnorm(60 * 3), ncol = 3),
                   matrix(rnorm(60 * 3, mean = 1.2), ncol = 3))
mon_ocd <- cpt_update(mon_ocd, stream_mv)
glance(cpt_delay(mon_ocd, truth = 61))
```

Two practical notes. The statistic reported is `ocd`'s *normalised*
statistic, already divided by its own threshold, so the comparison point
is 1 — printing the raw thresholds alongside it would show an alarm at
1.10 against a threshold of 16.4 and read as a bug. And the Monte Carlo
threshold calibration is the expensive part of building the monitor: a
minute or more at the default `patience = 5000`, which is why `mc_reps`
is lowered here. Pass `thresh` directly when you already have thresholds.

# 5. Delay is a function of shift size

The whole point of an online detector is that a bigger change is noticed
sooner. This is small enough to measure directly — ten replicates per
cell, median over replicates:

```{r delay-study}
set.seed(2026)
median_delay <- function(shift, method, reps = 10, n = 200) {
  d <- vapply(seq_len(reps), function(i) {
    s <- c(rnorm(n), rnorm(n, mean = shift))
    cpt_delay(cpt_replay(s, method = method), truth = n)$median_delay
  }, numeric(1))
  median(d, na.rm = TRUE)
}
grid <- expand.grid(shift = c(1, 2, 3),
                    method = if (has_cpm) c("edetector", "cpm") else "edetector",
                    stringsAsFactors = FALSE)
grid$median_delay <- mapply(median_delay, grid$shift, grid$method)
grid
```

Ten replicates is a noisy estimate of a median, and the table above will
wobble with the seed. A larger run of the same design — 150 replicates per
cell, repeated under two seeds — gives median detection delays of
**7, 3 and 2** observations for `edetector` at shifts of 1, 2 and 3
standard deviations, and **10, 5 and 4** for `cpm`. Two things are worth
taking from that. Delay falls steeply in the shift size, so a monitor
tuned on a large change will feel unusably slow on a small one. And the
gap between the two detectors is real but modest where the change is
small: two or three observations at 1 sigma, against inter-quartile
ranges of [5, 12] and [7, 15] that overlap almost entirely — while
`edetector` is clearly and consistently quicker once the change is
obvious.

# 6. Assumptions: the thresholds are for independent observations

Both calibrations — the e-detector's $1/\alpha$ bound and `cpm`'s `arl0`
— assume the in-control observations are independent. Under the null on
iid noise they hold up. Measured over 20 in-control streams of 2000
observations at the defaults, `cpm` raises **3.0** false alarms on average
against the 4 that `arl0 = 500` implies by construction, and `edetector`
**11.5**. The spread matters as much as the mean, and is the reason to
distrust a single stream: cpm's counts ranged from 0 to 7 across the 20
(sd 2.0) and the e-detector's from 1 to 19 (sd 5.2), so agreement to within
one alarm is not something one run can establish.

Serial dependence breaks both, and not gently. Autocorrelated noise
wanders; a detector that reads a wander as a level shift alarms on it.

```{r assumptions}
set.seed(3)
false_alarms <- function(method, gen, reps = 5, n = 400) {
  mean(vapply(seq_len(reps),
              function(i) nrow(alarms(cpt_replay(gen(n), method = method))),
              numeric(1)))
}
iid <- function(n) rnorm(n)
ar1 <- function(n) as.numeric(stats::arima.sim(list(ar = 0.7), n))
methods <- if (has_cpm) c("edetector", "cpm") else "edetector"
data.frame(
  method = methods,
  iid = vapply(methods, false_alarms, numeric(1), gen = iid),
  ar1_rho_0.7 = vapply(methods, false_alarms, numeric(1), gen = ar1),
  row.names = NULL
)
```

Both inflate badly at $\rho = 0.7$, `cpm` by roughly **tenfold** in a
larger run of the same comparison. Neither is broken — they are answering
the question they were calibrated for — but a nominal `arl0 = 500` on
autocorrelated data is not a 500-observation run length, and reporting it
as one overstates the evidence behind every alarm. The options, in
order of how much they ask of you:

1. **Pre-whiten.** Fit an AR model to the baseline and monitor the
   residuals. Cheap, and it restores the calibration when the model is
   about right.
2. **Aggregate.** Monitor block means rather than raw observations. The
   dependence between blocks is weaker; the cost is delay measured in
   blocks.
3. **Recalibrate empirically.** Simulate in-control streams with the
   dependence you actually have, and set `alpha` or `arl0` to the level
   that delivers the run length you want. This is the honest route when
   the noise model is not AR.

Two further assumptions, worth stating because they are easy to miss.
The baseline must be **clean**: `edetector` and `ocd` take their reference
mean and scale from it, so a change inside the baseline is inherited as
the in-control state and the real change afterwards may be invisible.
And `relearn` (20 observations by default) is not cosmetic. A real change
is *persistent*, so a monitor that restarts against the stale pre-change
baseline alarms again on the very next observation and keeps alarming for
the rest of the stream — reporting one change as hundreds, all but the
first of which `cpt_delay()` counts as false alarms. Set `relearn = 0`
only when you want to see every threshold crossing.

```{r relearn}
nrow(alarms(cpt_replay(stream, method = "edetector", relearn = 20)))
nrow(alarms(cpt_replay(stream, method = "edetector", relearn = 0)))
```

# 7. What to report

For an online analysis, the summary that belongs in a paper is
`glance(cpt_delay(...))`: how many changes there were, how many were
detected, the delay distribution, the false-alarm count and the realised
run length. A list of alarm times without the delay accounting is not an
evaluation, and a covering or F1 score computed against alarm times is a
category error. Delay against false alarms has been the currency of
sequential monitoring since @page1954continuous; the tooling here just
makes the accounting automatic.

```{r report}
glance(cpt_delay(rep_e, truth = 200))
```

# References
