---
title: "Benchmarks: Runtime, False Alarms and Coverage"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggchangepoint)
```

Fifty methods share one interface in this package, which makes it easy to
swap one for another and easy to forget that they do not cost the same
thing. This article answers three questions a reader choosing a method
has, from the measurements that ship with the package:

1. **How long does it take**, and at what series length does it stop being
   usable at all?
2. **How often does it fire when there is nothing there?**
3. **Do the intervals cover?**

## How these numbers were produced, and what they are not

The tables come from the `cpt_runtimes`, `cpt_null_sizes` and
`cpt_calibration` data sets, generated by the scripts in `data-raw/` for
this release on one Linux x86-64 machine (AMD EPYC) with R 4.4.1 and the
engine versions current then, and read here when the page is built. The
runtimes are **one fit per cell**: they answer "will it finish?", and are
not a basis for saying one engine is some factor faster than another.
Re-measure on your own hardware before quoting a time.

## 1. Runtime

Simulated Gaussian series with four mean changes of two noise standard
deviations (three coordinates for the multivariate engines), each fit in
its own R process after a warm-up call, with a two-minute cap (five at a
million). A length is attempted only when the previous one finished in
under ten seconds.

```{r runtime}
rt <- cpt_runtimes
rt$cell <- ifelse(rt$status == "ok", format(round(rt$seconds, 2), nsmall = 2),
                  rt$status)
wide <- reshape(as.data.frame(rt[, c("method", "n", "cell")]),
                idvar = "method", timevar = "n", direction = "wide")
names(wide) <- c("method", "1e3", "1e4", "1e5", "1e6")[seq_along(wide)]
wide[is.na(wide)] <- ""
fastest <- rt[rt$n == 1e6 & rt$status == "ok", ]
# Fastest at a million first, then the rest alphabetically.
rank_1e6 <- match(wide$method, fastest$method[order(fastest$seconds)])
wide <- wide[order(is.na(rank_1e6), rank_1e6, wide$method), ]
knitr::kable(wide, row.names = FALSE,
             caption = "Seconds per fit (blank: not attempted)")
```

`r nrow(fastest)` methods fit a million points inside the cap. The pruned
optimal-partitioning family and the single-change test are fast enough that
length is not a consideration; the random-interval searches cost more
because they fit many sub-intervals by construction, which is also why they
find short changes the single-pass methods miss. The methods that stop at
1,000 or 10,000 are quadratic or worse in the length (`ecp`, `bocpd`,
`segneigh`'s cost matrix), solve a harder problem (`cpop`'s continuous
piecewise-linear fit), or were written for the lengths their fields see
(`strucchange`, `taylor`). None is broken; they are methods for short
series, and choosing one for a long series is a modelling decision rather
than a performance problem to tune around. `cpt_methods()` carries each
method's measured `cost` and `max_n`, and `cpt_recommend(n = )` leaves out
the ones that did not finish at the length you give it.

## 2. False alarms on pure noise

Independent Gaussian noise with no changepoint anywhere, for the univariate
engines the runtime table shows finishing each length quickly: 50
replicates at 1,000, 10 at 10,000 and 6 at 100,000. Every changepoint
reported is a false alarm.

```{r null}
ns <- cpt_null_sizes
ns$cell <- paste0(format(ns$mean_fp, nsmall = 2), " (", ns$reps, ")")
wide_ns <- reshape(as.data.frame(ns[, c("method", "n", "cell")]),
                   idvar = "method", timevar = "n", direction = "wide")
names(wide_ns) <- c("method", "1e3", "1e4", "1e5")[seq_along(wide_ns)]
wide_ns[is.na(wide_ns)] <- ""
knitr::kable(wide_ns[order(wide_ns$method), ], row.names = FALSE,
             caption = "Mean false changepoints per series (replicates)")
```

Most engines raise none at their defaults. The rows that do are worth
reading correctly:

- **`segmented`** always reports the breakpoint it is asked for
  (`npsi = 1`), change or no change; test it with `cpt_test()` before
  reporting it.
- **`smuce`** and **`hsmuce`** default to `alpha = 0.5`, the level that is
  optimal for estimating the number of changes, and which by design
  overestimates it in up to half of all series. Pass a smaller `alpha` for
  a test.
- **`cpm`** is a sequential monitor used in batch mode: its threshold is a
  false-alarm rate per observation, so it reports about `n / arl0` false
  changepoints on any series. Since 0.6.0 the default `arl0` scales with the
  length (five times `n`, up to the 50,000 cpm ships), and the result
  records the expected count.
- **`np`** over-detects on Gaussian noise, and more so as the series grows.

Zero false alarms in a handful of replicates is consistent with a true rate
anywhere below a few tenths per series, so a zero here means "not obviously
anti-conservative" rather than "exact". And it is measured on independent
noise: under autocorrelation almost every one of those zeros becomes a
positive count (`cpt_noise_benchmark`, and the pitfalls article).

## 3. Interval coverage

The calibration article has the measured coverage of each provenance of
`cpt_confint()`:

```{r coverage}
subset(cpt_calibration, kind == "coverage")[, c("call", "nominal",
                                                 "realised", "mcse")]
```

Two provenances are not in it yet. **NSP's regions** hold a *global*
guarantee across all intervals and survive selection, so they are
conservative by design. **Posterior intervals** from `bcp` can be close to
vacuous at 95%: its posterior probability of a change is sharply peaked
but has diffuse tails, so pushing the credible level out sweeps in most of
the series. `cpt_confint()` warns when that happens, and its `source`
column says which provenance each interval has, because a posterior
credible interval and a confidence interval are not interchangeable.
