---
title: "Calibration: What the Package Promises, and What It Delivers"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggchangepoint)
cal <- cpt_calibration
pct <- function(v) paste0(format(round(100 * v, 1), nsmall = 1), "%")
row_of <- function(pattern) cal[grepl(pattern, cal$call, fixed = TRUE), ][1, ]
```

Every interval in this package has a level, every test a size, and every
monitor a false-alarm rate. A number like that is a promise, and a promise
nobody measures can be wrong for years while every call succeeds: before
0.5.0 the e-detector's run-length bound was off by a factor of two, and it
was found only because someone measured it.

This page is the measurement. Each row is one guarantee the documentation
states, checked by simulation, with the realised value and its Monte Carlo
standard error. The table ships with the package as `cpt_calibration` and
is regenerated for each release by `data-raw/calibration.R`.

## The table

```{r table}
show <- cal
rl <- show$kind == "run_length"
show$nominal <- ifelse(rl, as.character(round(show$nominal)),
                       pct(show$nominal))
show$realised <- ifelse(rl, as.character(round(show$realised)),
                        pct(show$realised))
show$mcse <- ifelse(rl, as.character(round(show$mcse)), pct(show$mcse))
show$kind <- sub("_", " ", show$kind)
knitr::kable(show[, c("call", "kind", "nominal", "realised", "mcse",
                      "reps")],
             col.names = c("Call", "Kind", "Promised", "Delivered",
                           "MC s.e.", "Replicates"))
```

## Reading it

**Location intervals.** Coverage is scored on one change of 1.5 noise
standard deviations after observation 100 of 200, conditional on a
changepoint landing within 20 positions of the truth. `strucchange`'s
interval covers `r pct(row_of("strucchange")$realised)` of the time and the
residual bootstrap of `cpt_confint()` `r pct(row_of("bootstrap")$realised)`,
both at a promised 95%. SMUCE's jump intervals are simultaneous over all
changepoints at level `1 - alpha`, and conservative for a single change:
`r pct(row_of("smuce")$realised)`.

**Tests of a date fixed in advance.** Every family of `cpt_test_at()`
holds its size to within about two Monte Carlo standard errors. The exact
Poisson and Fisher tests are conservative, as tests of discrete data are
(`r pct(row_of("poisson")$realised)` and `r pct(row_of("binomial")$realised)`),
and the windowed search, which pays for looking across eleven dates with a
permutation distribution of the maximum, delivers
`r pct(row_of("window")$realised)`.

**Tests for a change anywhere.** At `n = 200` the CUSUM test's Kolmogorov
limit gives `r pct(row_of("cpt_test_null(x)")$realised)` and the sup-F test
`r pct(row_of("supF")$realised)`; Pettitt's approximate p-value is
conservative, `r pct(row_of("pettitt")$realised)`.

**A test at a location the data chose.** Forced to report one changepoint
on pure noise, the naive test at it rejects
`r pct(row_of("amoc")$realised)` of the time at a nominal 5%. This is the
row that explains `selection_adjusted = FALSE`: the p-value is valid only
for a location chosen without the data, which is what `cpt_test_at()` is
for.

**Monitors.** `cpm`'s in-control average run length is
`r round(row_of("cpm")$realised)` against a promised
`r row_of("cpm")$nominal`. The e-detector promises a lower bound, an average
run length of at least `1 / alpha`, and delivers
`r round(row_of("edetector")$realised)` at `alpha = 0.01`.

## What is not here yet

`cpt_select()`'s choice of the number of changepoints, `cpt_metrics()`
against a reference implementation, and the intervals of the Bayesian
engines have no row, because they have not been measured. A guarantee
without a row is a claim, not a measurement.

## Regenerating it

```r
# from the package source, with a pinned R
MEASURE_CORES=8 Rscript data-raw/calibration.R
Rscript data-raw/build_measured_data.R
```

Each cell draws from its own random stream, so the rows are independent
measurements.
