---
title: "Choosing a Method: a Decision Tree and a Cheatsheet"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggchangepoint)
wired <- subset(cpt_methods(), status == "available")
fams <- cpt_families()
supports <- function(ci) unique(fams$method[fams$change_in == ci])
list_of <- function(v) paste0("`", v, "`", collapse = ", ")
```

"Which of these fifty do I want?" is the first question every new user has.
This page answers it three ways: a tree to walk, a cheatsheet to scan, and
`cpt_recommend()`, which answers from the same measurements in code. Every
list below is computed from `cpt_methods()` when the page is built, so it
cannot drift from the registry.

## Decide before you look

Pick the method from what you know about the problem, before you see what
each method says about your series. A method chosen because it found a
change carries no valid p-value, and that selection is the easiest mistake
to make with fifty methods behind one call.

## The tree

```
What arrives?
|
+-- data still arriving, and you want an alarm ---> cpt_monitor()
|                                                   (edetector, cpm, ocd)
+-- a formula: did a regression's relationship change?
|       ---> cpt_detect(y ~ x, data, method = "strucchange" | "segmented" | "fastcpd")
|
+-- several series that should change together ---> the multivariate methods
|
+-- one finished series. What changes?
    |
    +-- its level (the mean)
    |     is the noise independent?   yes --> see "A change in mean" below
    |                                 no  --> cpt_assumptions() says so; see
    |                                         "Autocorrelated noise"
    +-- its spread, or level and spread together --> change_in = "var" / "meanvar"
    +-- its slope, the level staying continuous   --> change_in = "slope"
    +-- its whole distribution                    --> change_in = "distribution"
    +-- counts, 0/1 outcomes, waiting times       --> family = "poisson" |
                                                      "binomial" | "exponential"
```

## A change in mean

`r length(supports("mean"))` methods detect a change in mean. Which of them
depends on three things.

**How long is the series?** The univariate ones measured to fit a
1,000,000-point series within five minutes (`cpt_runtimes`):

```{r long}
long <- subset(wired, max_n >= 1e6 & univariate &
                 method %in% supports("mean"))
long[order(long$method), c("method", "engine", "cost", "scale_invariant")]
```

**Do you need to say where it could be?** Methods returning a location
interval or a significance region: `r list_of(subset(wired, ci)$method)`.
Methods returning a posterior over locations:
`r list_of(subset(wired, posterior)$method)`.

**Is the noise on unit scale?** These were measured to give a different
answer on the same series in different units, so standardise first or
choose another: `r list_of(subset(wired, scale_invariant %in% FALSE)$method)`.

## Autocorrelated noise

Measured on AR(1) noise with a lag-one correlation of 0.7, two real changes
and 300 observations, the calls with the fewest false positives that still
find the changes (`cpt_noise_benchmark`):

```{r ar1}
ar <- subset(cpt_noise_benchmark, regime == "ar1" & fp <= 1)
head(ar[order(-ar$hits, ar$fp), c("call", "hits", "fp")], 6)
```

A general-purpose detector on the same data, for comparison:

```{r ar1-pelt}
subset(cpt_noise_benchmark, regime == "ar1" & method == "pelt" &
         setting == "default")[, c("call", "hits", "fp")]
```

## Counts, binary outcomes and waiting times

Methods that fit a Poisson cost: `r list_of(unique(subset(cpt_families(), family == "poisson")$method))`.
A binomial cost: `r list_of(unique(subset(cpt_families(), family == "binomial")$method))`.
An exponential cost: `r list_of(unique(subset(cpt_families(), family == "exponential")$method))`.
`cpt_families()` has the engine call each combination becomes.

## One change, or several?

Some designs return at most one changepoint, which is right for a
before-and-after question and wrong for everything else:
`r list_of(subset(wired, max_cp %in% 1L)$method)`. Two of them
(`segmented` and `mcp`) return exactly as many as they are asked for, one
by default, whether or not the data change: test the answer with
`cpt_test()` before reporting it.

## The cheatsheet

| I need | Start with |
| :- | :- |
| A default for one series, one kind of change | `cpt_detect(x)` (`pelt`, on standardised data) |
| Intervals for the locations | `cpt_detect(x, method = "smuce")`, `"strucchange"`, `"nsp"` |
| A posterior probability of change at each time | `method = "bcp"`, `"beast"`, `"bocpd"` |
| Robustness to outliers | `cpt_detect(x, method = "binsegrcpp", family = "l1")` |
| A change in slope | `cpt_detect(x, method = "cpop", change_in = "slope")` |
| Counts | `cpt_detect(x, method = "pelt", family = "poisson")` |
| 0/1 outcomes | `cpt_detect(x, method = "smuce", family = "binomial")` |
| A regression break | `cpt_detect(y ~ x, data = d, method = "strucchange")` |
| Several series at once | `cpt_detect(X, method = "inspect")`, `"geomcp"`, `"ecp"` |
| A long series (10^6) | `method = "pelt"`, `"fpop"`, `"binsegrcpp"`, `"mosum"` |
| An alarm as data arrive | `cpt_monitor("edetector", baseline = b)` |
| A test at a date fixed in advance | `cpt_test_at(x, when = date)` |
| To know whether to believe the answer | `cpt_assumptions(fit)`, `cpt_robustness(fit)` |

## The same answer in code

`cpt_recommend()` reads the same tables and scores every method on what you
tell it; it returns the call to run, the measured numbers behind the score,
and a reason:

```{r recommend}
cpt_recommend(noise = "autocorrelated", n = 5000, n_expected = 2)
```

Given a fit you already made, it fills in what the fit knows (length,
dimension, data type, and whether the residuals are autocorrelated):

```{r recommend-fit}
set.seed(1)
fit <- suppressWarnings(cpt_detect(as.numeric(arima.sim(list(ar = 0.7), 400)),
                                   method = "pelt"))
head(cpt_recommend(fit = fit)[, c("method", "call", "why")], 3)
```
