---
title: "Five Packages, One Series: the Nile"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.height = 3,
                      fig.width = 7, dpi = 72, message = FALSE)
library(ggchangepoint)
has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
all_there <- all(vapply(c("strucchange", "segmented", "bcp"), has,
                        logical(1)))
```

"Which changepoint package should I use?" is asked again and again, and
usually answered in prose. This page answers it in code: five widely used
changepoint packages on CRAN, run on one real series, each through its own
interface and then through one.

The series is `datasets::Nile`, the annual flow of the Nile at Aswan from
1871 to 1970, which ships with R. Its flow drops around the turn of the
century, and every textbook on structural change uses it.

```{r nile-plot, fig.alt = "Annual Nile flow, 1871 to 1970"}
autoplot(cpt_detect(Nile, method = "strucchange"), show_ci = TRUE)
```

## Each package in its own words

Five interfaces, five return types, three conventions for what "the
changepoint" means:

```{r native, eval = all_there}
y <- as.numeric(Nile)

# changepoint: an S4 object; cpts() is the last index of each segment
changepoint::cpts(changepoint::cpt.meanvar(y, method = "PELT"))

# strucchange: a formula, and breakpoints() with a confint() method
bp <- strucchange::breakpoints(Nile ~ 1)
bp$breakpoints
confint(bp)$confint

# segmented: a fitted lm, and a breakpoint on the covariate's own scale
d <- data.frame(flow = y, year = as.numeric(time(Nile)))
segmented::segmented(lm(flow ~ year, data = d), seg.Z = ~ year)$psi

# bcp: a posterior probability of a change at each time
set.seed(1)
which.max(bcp::bcp(y)$posterior.prob)

# ecp: a matrix in, and estimates that are the FIRST index of each segment
set.seed(1)
ecp::e.divisive(matrix(y), min.size = 2)$estimates
```

`changepoint` and `strucchange` report the last year of the old regime,
`ecp` the first year of the new one, `bcp` a probability per year and
`segmented` a point on the covariate, estimated on a continuous scale. The
same answer, observation 28 (1898), is written four ways.

## The same five through one interface

```{r unified, eval = all_there}
calls <- c(pelt = "meanvar", strucchange = "mean", segmented = "slope",
           bcp = "mean", ecp = "distribution")
fits <- Map(function(m, ci) {
  suppressWarnings(cpt_detect(Nile, method = m, change_in = ci, seed = 1))
}, names(calls), calls)
do.call(rbind, lapply(names(fits), function(m) {
  cp <- fits[[m]]$changepoints
  data.frame(method = m, engine = fits[[m]]$versions$engine,
             change_in = calls[[m]],
             years = paste(cp$cp_index, collapse = ", "),
             interval = if ("ci_lower" %in% names(cp)) {
               paste(time(Nile)[cp$ci_lower], time(Nile)[cp$ci_upper],
                     sep = " to ", collapse = "; ")
             } else "")
}))
```

Every result is the same `ggcpt` object in the same convention (the last
observation before the change), with the years attached, so the table is
one `rbind()` rather than five extraction recipes.

## What the table says

- **Four of five find 1898.** `strucchange` adds a 95% interval, 1895 to
  1902, and `bcp`'s posterior probability of a change peaks there too.
- **`pelt` with a change in mean and variance finds two more**, in 1874 and
  1876: short segments of a few years each, which a variance model can fit
  closely. A change in mean alone, on the raw flows, is worse (the
  [pitfalls article](pitfalls.html) shows 42 changepoints), because the
  Normal cost assumes the noise has standard deviation 1.
- **`segmented` answers a different question.** It fits a broken line, a
  change in slope with the level continuous, so a step appears as a kink,
  here near 1913. It is the right tool for dose-response and trend breaks,
  and the wrong one for a level shift.

## The next questions, in the same grammar

```{r next}
fit <- cpt_detect(Nile, method = "strucchange")
cpt_effect(fit)[, c("cp_index", "before", "after", "delta", "delta_lower",
                    "delta_upper", "pct_change")]
cpt_test_at(Nile, when = 1899)[, c("cp_index", "estimate", "p_value",
                                   "method")]
cpt_assumptions(fit)
```

The flow fell by about a fifth. Had the date been fixed in advance (the
first Aswan dam was built between 1898 and 1902), `cpt_test_at()` is an
exact test with no selection to adjust for; a test at the detected date,
chosen by the same data, is not.
